# Python interoperability {#sec-bkg-python-interoperability}

{{< include _start.qmd >}}

## Introduction

Methods discussed in this book are usually available as either R or Python packages. Ideally, users should be able to leverage the full range of available tools for their analyses. Methods should be selected based on scientific merit (ideally demonstrated by neutral benchmarks), and independent of having been implemented in a given programming language (R or Python) or framework (e.g. Bioconductor or `Seurat`).

[R's application to statistical analysis of spatial data dates back decades, primarily in epidemiological and geospatial research. As a result, various tools for spatial analyses have been established. For example, `r BiocStyle::CRANpkg("sp")` provides a "coherent set of classes and methods for \[...\] points, lines, polygons, and grids"; `r BiocStyle::CRANpkg("spatstat")` and, more recently, `r BiocStyle::CRANpkg("sf")` provide tools for spatial point pattern and vector data, respectively.]{.aside} For single-cell and spatial omics data analysis, being able to leverage different ecosystems is especially powerful. On the one hand, Python offers superb infrastructure for image analysis and machine learning-based approaches. On the other hand, the R programming language has been historically dedicated to statistical computing; as a result, many modern R methods for spatial omics data build on a solid foundation of tools for spatial statistics and statistical modeling in general.

Different data structures, although standardized within a given framework, make switching between languages and tools somewhat cumbersome. In the realm of single-cell and spatial omics, all Bioconductor tools are built around `r BiocStyle::Biocpkg("SummarizedExperiment")`-derived classes, while [Seurat](https://satijalab.org/seurat/) [@Hao2023-Seurat], [Giotto](https://giottosuite.com/) [@Chen2025-Giotto-Suite], and `r BiocStyle::Githubpkg("BIMSBbioinfo/VoltRon")` each rely on their own object definitions. In Python, [Scanpy](https://scanpy.readthedocs.io/) [@Wolf2018-Scanpy] and [Squidpy](http://squidpy.readthedocs.io/) [@Palla2022-Squidpy] use [AnnData](https://anndata.readthedocs.io/). Attempts to alleviate the problem are being made -- e.g. `r BiocStyle::Biocpkg("zellkonverter")`, `r BiocStyle::Biocpkg("anndataR")` [@Deconinck2026-anndataR], and functions from `Seurat` allow for conversion between Python's `AnnData` and R/Bioconductor's `r BiocStyle::Biocpkg("SingleCellExperiment")` or `r BiocStyle::Biocpkg("SpatialExperiment")`.

[[Quarto](https://quarto.org/) is the successor to [R Markdown](https://rmarkdown.rstudio.com/) (by Posit, formerly known as RStudio). Similar to *.Rmd* files, *.qmd* files can include scientific content (e.g. cross-referencing, LaTeX-based equations), and can be published in multiple output formats (HTML, PDF, etc.). This book is built using Quarto.]{.aside} On a higher level, tools that enable interoperability between programming languages have become available. For example, [reticulate](https://rstudio.github.io/reticulate/) provides an R interface to Python, including support to translate between objects from both languages; `r BiocStyle::Biocpkg("basilisk")` facilitates Python environment management within the Bioconductor ecosystem, and can also be interfaced with `reticulate`; and Quarto can generate dynamic reports from code in different languages.

In this chapter, we demonstrate examples showing how to set up Python environments and interact with `anndata` objects in R, using either a virtual or Conda environment, together with [reticulate](https://rstudio.github.io/reticulate/) or `r BiocStyle::Biocpkg("basilisk")` and `r BiocStyle::Biocpkg("anndataR")`.

## Dependencies

```{r libraries, message=FALSE, warning=FALSE}
#| code-fold: false
library(anndataR)
library(basilisk)
library(jsonlite)
library(VisiumIO)
library(OSTA.data)
library(reticulate)
library(SpatialExperiment)
```

## Python environment and reticulate

We here set up a virtual environment to interact with Python `anndata` objects in R. 
In order to build this book programmatically, we install Python and pip dependencies 
*from scratch* using [`reticulate`](https://rstudio.github.io/reticulate/).

However, if you are following this example on a laptop or another system, 
it may be easier to specify an existing Python installation to use instead. 
This will avoid cluttering your system with additional Python installations; 
see the below section to, e.g., activate an existing Conda environment.

### Set up virtual environment

First, we create a virtual environment, installing Python dependencies
required in this chapter, as well as @sec-img-cell-cell-communication.

```{r venv-, echo=FALSE, eval=FALSE}
# we could create the venv using 'requirements', but
# this would require installing OSTA in its entirety
# to make package files accessible via 'system.file'
install_python(version=ver <- "3.8.18")
txt <- system.file("extdata", "requirements.txt", package="OSTA")
virtualenv_create(envname=env <- "OSTA", python=ver, requirements=txt)
use_virtualenv(env, required=TRUE)
```

```{r req, echo=FALSE, eval=FALSE}
# instead, we construct a character vector
# to be used in the following code chunk
dir <- dirname(getwd())
txt <- file.path(dir, "extdata", "requirements.txt")
req <- read.delim(txt, header=FALSE)[[1]]
req <- paste0("'", req, "'", collapse=", ")
```

```{r venv, message=FALSE, results="hide"}
req <- c('access==1.1.9', 'affine==2.4.0', 'anndata==0.9.2', 'anndata2ri==1.3.1', 'attrs==23.1.0', 'backports.zoneinfo==0.2.1', 'beautifulsoup4==4.12.2', 'certifi==2023.11.17', 'cffi==1.16.0', 'charset-normalizer==3.3.2', 'click==8.1.7', 'click-plugins==1.1.1', 'cligj==0.7.2', 'commot==0.0.3', 'contourpy==1.1.1', 'cycler==0.12.1', 'decorator==4.4.2', 'deprecation==2.1.0', 'esda==2.4.3', 'fiona==1.9.5', 'fonttools==4.45.1', 'gensim==4.3.2', 'geopandas==0.13.2', 'get-annotations==0.1.2', 'giddy==2.3.4', 'h5py==3.10.0', 'idna==3.4', 'igraph==0.10.8', 'importlib-metadata==6.8.0', 'importlib-resources==6.1.1', 'inequality==1.0.0', 'Jinja2==3.1.2', 'joblib==1.3.2', 'karateclub==1.3.3', 'kiwisolver==1.4.5', 'leidenalg==0.10.1', 'Levenshtein==0.23.0', 'libpysal==4.7.0', 'llvmlite==0.41.1', 'mapclassify==2.6.0', 'MarkupSafe==2.1.3', 'matplotlib==3.7.4', 'mgwr==2.1.2', 'momepy==0.6.0', 'mpmath==1.3.0', 'natsort==8.4.0', 'networkx==2.6.3', 'numba==0.58.1', 'numpy==1.22.4', 'packaging==23.2', 'pandas==1.3.5', 'patsy==0.5.3', 'Pillow==10.1.0', 'platformdirs==4.0.0', 'plotly==5.18.0', 'pointpats==2.4.0', 'POT==0.9.1', 'PuLP==2.7.0', 'pycparser==2.21', 'pygeos==0.14', 'PyGSP==0.5.1', 'pynndescent==0.5.11', 'pyparsing==3.1.1', 'pyproj==3.5.0', 'pysal==23.7', 'python-dateutil==2.8.2', 'python-igraph==0.10.8', 'python-Levenshtein==0.23.0', 'python-louvain==0.16', 'pytz==2023.3.post1', 'quantecon==0.7.1', 'rapidfuzz==3.5.2', 'rasterio==1.3.9', 'rasterstats==0.19.0', 'requests==2.31.0', 'rpy2==3.5.14', 'Rtree==1.1.0', 'scanpy==1.9.6', 'scikit-learn==1.3.2', 'scipy==1.10.1', 'seaborn==0.12.2', 'segregation==2.5', 'session-info==1.0.0', 'setuptools==68.2.2', 'shapely==2.0.2', 'simplejson==3.19.2', 'six==1.16.0', 'smart-open==6.4.0', 'snuggs==1.4.7', 'soupsieve==2.5', 'spaghetti==1.7.4', 'spglm==1.0.8', 'spint==1.0.7', 'splot==1.1.5.post1', 'spopt==0.5.0', 'spreg==1.4', 'spvcm==0.3.0', 'statsmodels==0.14.0', 'stdlib-list==0.10.0', 'sympy==1.12', 'tenacity==8.2.3', 'texttable==1.7.0', 'threadpoolctl==3.2.0', 'tobler==0.9.0', 'tqdm==4.66.1', 'tzlocal==5.2', 'umap-learn==0.5.5', 'urllib3==2.1.0', 'zipp==3.17.0')
install_python(version=ver <- "3.8.18")
virtualenv_create(envname=env <- "OSTA", python=ver, packages=req)
use_virtualenv(env, required=TRUE)
```

### Using Conda

Alternatively, an existing Conda environment -- including the corresponding 
Python installation and software dependencies -- could be used as follows:

```{r conda, eval=FALSE}
# path to Conda binary
bin <- ".../bin/conda" 
options(reticulate.conda_binary=bin)
# activate environment
use_condaenv("<name>")
```

### Using `basilisk`

Alternatively, we can use `r BiocStyle::Biocpkg("basilisk")` to install and manage Python environments. This is an alternative to setting up a Conda environment with an existing Python installation. Using `basilisk` provides a self-contained environment, which helps avoid problems due to interactions with other environments on your system, and provides a higher level of reproducibility. This is especially useful for package development by advanced users. However, setup may be more challenging in some cases. A minimal example is provided here for completeness:

```{r basilisk, eval=FALSE}
# set up environment using 'basilisk'
env <- BasiliskEnvironment(
    pkgname="base", 
    envname="basilisk", 
    pip="numpy==2.4.0",
    packages="python=3.12")
# activate virtual environment
use_virtualenv(obtainEnvironmentPath(env))
```

## Interfacing between languages

### SingleCellExperiment

#### Calling Python

After configuring Python, R commands can now be run using `reticulate`; 
for more details on syntax, see [here](https://rstudio.github.io/reticulate/). 
[Note that running code in this way comes with a small overhead of starting up 
a Python session in the background. But this is typically small compared to the 
runtime required to run computation-heavy methods, or when analyzing large-scale
single-cell and spatial data (hundreds of thousands of cells or more).]{.aside}

```{python load-ad, message=FALSE, warning=FALSE}
import scanpy
id = "V1_Mouse_Brain_Sagittal_Posterior"
ad = scanpy.datasets.visium_sge(sample_id=id)
ad.var_names_make_unique()
ad
```

#### Continuing in R

We can access any of the variables above in R.
For basic outputs, this works out of the box:

```{r show-ad}
range(py$ad$obs$array_row)
```

`reticulate` also supports a few direct [type conversions](https://rstudio.github.io/reticulate/#type-conversions) (e.g. dictionary $\leftrightarrow$ named list). 
In the example demonstrated here, we use `anndataR` to convert from `AnnData` to `SingleCellExperiment`:

```{r ad2sce, message=FALSE, warning=FALSE}
(sce <- py$ad$as_SingleCellExperiment(x_mapping="counts"))
```

#### Back to Python

We can also do the reverse, i.e., go from R's `SingleCellExperiment` to Python's `AnnData`:

```{r sce2ad}
(ad <- as_AnnData(sce, x_mapping="counts"))
```

### SpatialExperiment

Since the `SpatialExperiment` class extends `SingleCellExperiment` (see @sec-bkg-infrastructure), 
conversion operations discussed above are also applicable to `SpatialExperiment`. 
However, to accomplish a full conversion from the `AnnData` object, 
we need to manually insert the spatial information using `reticulate` directly.

#### Starting in R

For this use case with `SpatialExperiment`, we will use the dataset from @Janesick2023-high-res, which includes Visium measurements on human breast cancer tissue.

```{r load-spe, message=FALSE, warning=FALSE, results="hide"}
id <- "Visium_HumanBreast_Janesick"
pa <- OSTA.data_load(id)
dir.create(td <- tempfile())
unzip(pa, exdir = td)
obj <- TENxVisium(
    spacerangerOut=file.path(td, "outs"), 
    format="h5", 
    images="lowres")
(spe <- VisiumIO::import(obj))
```

We also need to parse the original scaling information (i.e., scale factor) 
for spots and images available in the standard Visium output. 
We will use this later during conversion.

```{r sfs, message=FALSE, results="hide"}
fnm <- "scalefactors_json.json"
fnm <- file.path(td, "outs", "spatial", fnm)
sfs <- read_json(fnm)
```

We again use the `as_AnnData` function from `anndataR` from the previous example, and convert the `SingleCellExperiment`-relevant components of the `SpatialExperiment` object to an `AnnData` object.

```{r spe2ad, message=FALSE}
mtx <- as(assay(spe), "dgCMatrix")
assay(spe) <- mtx
(ad <- as_AnnData(spe))
```

We can now populate the `uns` and `obsm` components of the `AnnData` object 
with spatial coordinates and images. We start with the coordinates.

```{r xy}
spatialCoordsNames(spe) <- c("x", "y")
obsm <- list(spatial=spatialCoords(spe))
```

Now, let's create the `uns` component. The list of `uns` should be composed of as many samples as the images in the `SpatialExperiment` object. Also, each sample entry in the list should have two elements, one for the image and the other for the scaling information.

```{r uns}
# get image metadata
imgdata <- imgData(spe)

# get image
img <- imgRaster(spe)
img <- apply(img, c(1, 2), \(x) col2rgb(x))
img <- aperm(img, perm=c(2, 3, 1))
img <- img / 255

# create uns
uns <- list(images=list(lowres=img), scalefactors=sfs)
uns <- list(spatial=setNames(list(uns), imgdata$sample_id))
```

Now let's insert the components to the `AnnData` object, 
and write back to an `.h5ad` file.

```{r write-h5ad, message=FALSE, warning=FALSE}
ad$obs$library_id <- imgdata$sample_id
ad$obsm <- obsm
ad$uns <- uns
ad$write_h5ad("spe.h5ad")
```

#### Calling Python

Now that we have converted the `SpatialExperiment` object to `AnnData` format, 
we can run Python code per usual (e.g., continue analysis using `squidpy`).
As a proof of concept, we here visualize an exemplary gene's counts spatially:

```{python plot-ad, fig.width=4.5, fig.height=4.5}
import anndata
import matplotlib
import matplotlib.pyplot as plt

ad = anndata.read_h5ad("spe.h5ad")
ad.var_names = ad.var["Symbol"].astype(str)

xy = ad.obsm["spatial"]
z = ad[:,"ERBB2"].layers["counts"].toarray()

plt.scatter(xy[:,0], xy[:,1], c=z, s=5, cmap="turbo")
plt.gca().set_aspect("equal")
plt.title("ERBB2")
```

```{r cleanup-h5ad, include=FALSE}
# clean up by removing the `spe.h5ad` file
file.remove("spe.h5ad")
```

## Appendix

::: {.callout-tip icon="false" title="Further reading"}

- This [single-cell best practices](https://www.sc-best-practices.org/introduction/interoperability.html)
Python book chapter covers interoperability from Python/scverse's view, 
including H5AD file handling plus a few general remarks on file formats, 
and disk-based vs. in-memory interoperability.

- `r BiocStyle::Biocpkg("anndataR")` [@Deconinck2026-anndataR]: 
R package and community project to work with `AnnData` objects in R, including 
conversion to and from `SingleCell/SpatialExperiment` and `Seurat` objects.

- [reticulate](https://rstudio.github.io/reticulate/) [@Ushey2017-reticulate]: 
R package and framework to call Python and run Python code in R, translate 
between R and Python objects, and manage virtual and Conda environments.

- `r BiocStyle::Biocpkg("basilisk")` [@Lun2022-basilisk]: 
R package to install and manage Python environments in R packages and sessions.

:::

### References {.unnumbered}

{{< include _stop.qmd >}}
