# Ecosystem {#sec-bkg-ecosystem}

{{< include _start.qmd >}}

## Introduction

This chapter provides additional background about the Bioconductor ecosystem
by exploring the space of packages that revolve around spatial and single cell
omics data. We deliberately include the latter, since many tools developed for
single cell data are either directly applicable to, or lay the foundation for 
tools developed for spatial data.

Here, we will rely on `r BiocStyle::Biocpkg("biocViews")` [@Carey2006-BiocViews] 
that are provided in the `DESCRIPTION` of each package. These may be browsed 
online -- in a non-programmatic manner -- on the [Bioconductor BiocViews 
website](https://www.bioconductor.org/packages/release/BiocViews.html).

::: aside
Note that the data reported herein may be inaccurate, since packages might
provide inaccurate/incomplete `biocViews`. Especially the "Spatial" term relied 
upon here was added fairly recently and might be missing from older packages.
:::

### Exploratory

The `r BiocStyle::Biocpkg("BiocPkgTools")` package [@Su2018-BiocPkgTools]
provides tools to access and explore Bioconductor package metadata in R, 
including their details (e.g., authors) and (monthly) download statistics.

`biocExplore()` provides an interactive way to explore the package space:

```{r load-lib, message=FALSE}
library(BiocPkgTools)
foo <- tryCatch(
    biocExplore(), 
    error=\(e) NULL) # protext against unstable connection
```

`biocPkgList()` retrieves the full Bioconductor software package listing
(at the time of calling the function), including associated metadata.
These include `biocViews`, which we can use to identify packages of interest:

```{r load-dat, message=FALSE, echo=-1}
options(width = 100)
# retrieve current package record
df <- biocPkgList()
# helper function to get indices of packages
# that contain 'biocViews' specified by 'x'
.f <- \(x, y=df) vapply(y$biocViews, \(.) all(x %in% .), logical(1))
# view "Spatial" packages
df$Package[.f("Spatial")]
```

We can also browse for packages that also include more specific terms, for example:

```{r filt-dat}
# view "Spatial Clustering" packages
df$Package[.f(c("Spatial", "Clustering"))]
```

### Metrics

We will now have a look at the package ecosystem in a more quantitative manner.
Specifically, we will investigate how the number of available packages evolves
over time, and how long individual packages remain available (their "lifetime").
Lastly, we will quantify the number of times more specific subterms (co-)occur.

#### Number of packages

Let's first summarize the number of packages available over time,
focusing specifically on "SingleCell" and "Spatial" packages.

::: aside
Note that we here rely on the first and last date at which a package was
available through Bioconductor; some packages might have been deprecated
at one point or another.
:::

```{r numb-pkg-prep, message=FALSE, results="hide"}
#| code-fold: true
# dependencies
library(dplyr)
library(ggplot2)
# specify 'biocViews' of interest
names(ids) <- ids <- c("SingleCell", "Spatial")
now <- as.Date(format(Sys.Date(), "%Y-%m-%d"))
gg <- lapply(ids, \(id) {
    # get metadata & simplify naming
    nm <- df$Package[.f(id)]
    ys <- getPkgYearsInBioc(nm) |>
        mutate(first=first_version_release_date) |>
        mutate(last=last_version_release_date) |>
        mutate(last=case_when(is.na(last)~now, TRUE~last)) |>
        filter(!is.na(first))
    # complete months between first/last dates
    ys <- lapply(split(ys, ys$package), \(.) {
        data.frame(
            package=.$package, 
            date=seq(.$first, .$last))
    }) |> do.call(what=rbind) 
    # get cumulative number of packages available each month
    ys |> group_by(date) |> count()
}) |> bind_rows(.id="biocViews")
```

```{r numb-pkg-plot, message=FALSE, fig.width=5, fig.height=4}
#| code-fold: true
ggplot(gg, aes(date, n, col=biocViews)) + 
    geom_line(linewidth=0.8) +
    geom_smooth(data=filter(gg, n >= 5), 
        method="lm", se=FALSE, linewidth=1) +
    scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
    labs(x=NULL, y="# packages") +
    theme_bw() + theme(
        aspect.ratio=1, 
        panel.grid.minor=element_blank(),
        axis.text.x=element_text(angle=45, hjust=1))
```

To get an idea of the project's "growth" in this space, we can regress the 
number of packages against time using a linear model (LM). We here skip 
dates at which very few packages were available/being developed:

```{r perc-pgk, echo=-c(3, 4)}
#| code-fold: true
# fit linear model where x = 'date', y = # packages
# (starting at 'date' where at least 5 packages exist)
.gg <- filter(gg, n >= 5)
# get shift in years between "SingleCell" and "Spatial"
dy <- round(as.integer(diff(slice_min(group_by(.gg, biocViews), date)$date))/365, 2)
(bs <- .gg |>
    group_by(biocViews) |>
    group_split() |>
    # return coefficients = # packages 
    # added each month (on average)
    sapply(\(.) coef(lm(n~date, .))[[2]]) |>
    setNames(sort(unique(gg$biocViews))))
```

The resulting coefficients tell us that about `r round(bs["SingleCell"]*12, 2)` 
single cell and `r round(bs["Spatial"]*12, 2)` spatial packages are being added
each year (on average), and there is a delay of about `r dy` years between them.

#### Package lifetimes

We can further inspect the "lifetime" of packages, i.e., how long they remain 
(installable) on Bioconductor:

::: aside
A package will be deprecated if it fails to build and/or pass checks on the 
Bioconductor Build System (BBS), provided its maintainer is unresponsive and 
does not take action to fix the package before the next (six-monthly) release. 
:::

```{r pkg-life, results="hold", message=FALSE, fig.width=5, fig.height=4}
#| code-fold: true
gg <- lapply(ids, \(id) {
    nm <- df$Package[.f(id)]
    ys <- BiocPkgTools:::getPkgYearsInBioc(nm)
}) |> bind_rows(.id="biocViews") |>
    mutate(years=approx_years_in) |>
    filter(!is.na(years))
mu <- gg |>
    group_by(biocViews) |>
    summarise_at("years", mean)
# print
cat("years in Bioconductor:\n")
summary(gg$approx_years_in)
# plot
ggplot(gg, aes(years, fill=biocViews)) + 
    geom_histogram(alpha=1/3, binwidth=0.5) +
    geom_vline(
        linewidth=1, data=mu,
        aes(xintercept=years, col=biocViews)) +
    scale_x_continuous(breaks=seq(0, 100, 2)) +
    labs(x="lifetime (years)", y="# packages") +
    theme_bw() + theme(
        aspect.ratio=1, 
        panel.grid.minor=element_blank(),
        axis.text.x=element_text(angle=45, hjust=1))
```

#### Subterms

Finally, let's investigate more specific `biocViews`, e.g., the type of work tasks 
different packages cover. We will first count the number of packages that list
specific terms that might be of interest in the context of single cell and
spatial omics data analysis. Secondly, we will visualize their co-occurrence.

::: aside

Note that packages can list an arbitrary number of terms so that packages can be
counted in more than one category; e.g., "Visualization" appears in *most*.

:::

```{r pkg-catego-one, fig.width=5, fig.height=4}
#| code-fold: true
i <- c("SingleCell", "Spatial")
j <- c(
    "BatchEffect", "Normalization", "QualityControl", "Visualization", # WorkflowStep
    "Clustering", "DimensionReduction", "FeatureExtraction", # StatisticalMethod
    "DifferentialExpression", "GeneSetEnrichment") # BiologicalQuestion
names(i) <- i; names(j) <- j
# count packages for each pair of terms
gg <- mapply(
    i=i, j=rep(j, each=2), 
    SIMPLIFY=FALSE, \(i, j) {
        n <- sum(.f(c(i, j)))
        data.frame(i, j, n)
    }) |> do.call(what=rbind)
# order x-axis by total
xo <- gg |>
    group_by(j) |>
    summarise_at("n", sum) |>
    arrange(n) |>
    pull(j)
ggplot(gg, aes(j, n, fill=i)) + 
    scale_x_discrete(limits=xo) +
    geom_col(position="dodge", alpha=2/3) +
    labs(x=NULL, y="# packages", fill="biocViews") +
    theme_bw() + theme(
        aspect.ratio=1, 
        panel.grid.minor=element_blank(),
        axis.text.x=element_text(angle=45, hjust=1))
```

Because there are yet very few "Spatial" packages, we will pool "Spatial" and
"SingleCell" tools when counting the number of times different terms appear
together; i.e., `biocViews` contain "Spatial" OR "SingleCell", together with
"Clustering" AND "Visualization", etc. (For concise labeling, we abbreviate
`biocViews` to capital letters only; e.g., "BatchEffects" becomes "BE".)

```{r pkg-catego-two, fig.width=5, fig.height=4}
#| code-fold: true
gg <- lapply(i, \(i) {
    lapply(seq_along(j), \(n) {
        js <- combn(j, n, simplify=FALSE)
        lapply(js, \(j) {
            n <- sum(.f(c(i, j)))
            j <- gsub("[a-z]", "", j)
            j <- paste(j, collapse="+")
            data.frame(i, j, n)
        }) |> do.call(what=rbind)
    }) |> do.call(what=rbind)
}) |> do.call(what=rbind) |>
    group_by(j) |>
    summarize_at("n", sum) |>
    slice_max(n, n=30)
yo <- gg$j[order(gg$n)]
ggplot(gg, aes(n, j)) + 
    geom_col(alpha=1/3, fill="blue") +
    labs(y=NULL, x="# packages") +
    scale_y_discrete(limits=yo) +
    theme_bw() + theme(
        aspect.ratio=1, 
        panel.grid.minor=element_blank(),
        axis.text.x=element_text(angle=45, hjust=1))
```

## Appendix

::: {.callout-tip icon="false" title="Further reading"}

- The **Bioconductor project** has been described approximately once per decade,
summarizing the project's current status, challenges, and future perspectives
[@Gentleman2004-Bioconductor; @Huber2015-Bioconductor; @Carey2025-Bioconductor].

- Many additional e-books, resources, and publications capture additional 
aspects of the project, such as educational efforts [@Drnevich2025-bioc-edu];
see [here](https://www.bioconductor.org/help/publications/) for a full list.

- The **scverse project** was first describes in 2023 and, arguably inspired
by R/Bioconductor, set out to build a unified computational ecosystem for 
single-cell and spatial omics data analysis in Python [@Virshup2023-scverse].

:::

### References {.unnumbered}

{{< include _stop.qmd >}}
