# Spatial Statistics

## Preamble

### Introduction

> Start with some words on what motivates this chapter.

### Dependencies

```{r loading, message = FALSE, warning = FALSE}
library("SpatialFeatureExperiment")
library("SFEData")
library("Voyager")
library("ggplot2")
library("scran")
library("BiocParallel")
library("dplyr")
library("spdep")
library("openxlsx")
library("spatialFDA")
library("spatstat")

bp <- MulticoreParam(10)
set.seed(77)

sfe <- JanesickBreastData()

# load the official 10X annotations 
labels <- read.xlsx("https://cdn.10xgenomics.com/raw/upload/v1695234604/Xenium%20Preview%20Data/Cell_Barcode_Type_Matrices.xlsx", sheet = 4)

# add cell id to label dataframe
labels$Barcode <- rownames(labels)
# add the cell type labels to the spe
matchedDf <- as.data.frame(colData(sfe)) |>
  left_join(as.data.frame(labels), by = join_by("Barcode"))

colData(sfe) <- DataFrame(matchedDf)

sfe <- sfe[, colSums(counts(sfe)) > 0]
sfe <- sfe[, !is.na(colData(sfe)[["Cluster"]])]
sfe <- logNormCounts(sfe)

xy <- spatialCoords(sfe)
# basic theme for spatial plots
theme_xy <- list(
    coord_equal(expand=FALSE), 
    theme_void(), theme(
        plot.margin=margin(l=5),
        legend.key=element_blank(),
        panel.background=element_rect(fill="black")))
```


{{< include crs-spat-stat-lattice.qmd >}}

{{< include crs-spat-stat-pp.qmd >}}

> To keep things concise, use a collapsed box when you see fit (e.g., more details, advanced content, etc.):

::: {.callout-note collapse="true" title="itbeme"}

Anything can go in here (text/math, code chunks)...

```r
# do some fancy things
```

:::

## Appendix

### Saving {.unnumbered}

> If applicable, save objects for downstream.

```{r saving upstream}
#saveRDS(obj, "xxx-foo.rds")
```

### Benchmarks {.unnumbered}

> Bullet list of benchmark papers, ideally giving a 1-2 sentence summary of what's being compared (e.g., how many methods/datasets, which technologies, etc.).

### References {.unnumbered}

> Auto-populated; add nothing.