# Quality control {#sec-img-quality-control}

{{< include _start.qmd >}}

## Preamble

### Introduction

The analyses demonstrated here will use 313-plex Xenium data (10x Genomics) 
of a human breast cancer biopsy section [@Janesick2023-high-res], and
1k-plex CosMx data (Bruker) of a mouse brain section.
These datasets can be accessed through Bioconductor's `ExperimentHub` packages:
`r BiocStyle::Biocpkg("STexampleData")` and `r BiocStyle::Biocpkg("OSTA.data")`.

::: {.aside}
The CosMx dataset includes additional information, such as fluorescent stains, 
cell boundaries, field of view placement, etc. that allow us to demonstrate 
more intricate quality control metrics using the `SpaceTroper` package.
:::

Throughout this chapter, we will switch between both Xenium (`xen` object) 
and CosMx (`cos` object) datasets to showcase quality control (QC) metrics 
that are standard for scRNA-seq data (using `r BiocStyle::Biocpkg("scuttle")`), 
and ones that are more specific to some imaging-based ST data types
(using `r BiocStyle::Biocpkg("SpaceTrooper")` [@Banzi2025-SpaceTrooper]).
Lastly, we will save the post-QC Xenium dataset for use in later chapters.

### Dependencies

```{r deps, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
library(ggspavis)
library(OSTA.data)
library(patchwork)
library(scrapper)
library(scuttle)
library(SpaceTrooper)
library(SpatialExperiment)
library(SpatialExperimentIO)
library(STexampleData)
library(tidyr)
```

## Importing data

`r BiocStyle::Biocpkg("SpatialExperimentIO")` provides functions to import 
CosMx (Bruker), Xenium (10x Genomics), and MERSCOPE (Vizgen) datasets into
`SingleCell/SpatialExperiment` objects, starting from the outputs provided 
by the three providers. 10x Genomics-specific readers are provided by 
`r BiocStyle::Biocpkg("TENxIO")`, `r BiocStyle::Biocpkg("VisiumIO")`, and 
`r BiocStyle::Biocpkg("XeniumIO")`; data from most commercial platforms can 
also be read using `r BiocStyle::Biocpkg("SpatialFeatureExperiment")` functions.

::: {.aside}
See [@sec-bkg-importing-data] for details on data importing and representation.
:::

### From Xenium

We load the Xenium dataset from `r BiocStyle::Biocpkg("STexampleData")`:

```{r load-data-xen, message=FALSE, results="hold"}
(xen <- Janesick_breastCancer_Xenium_rep1())
# required for visuals
xen$in_tissue <- TRUE 
xen$sample_id <- "Xenium"
```

### From CosMx

To load the CosMx data into a `SpatialExperiment` object, 
we use `SpatialExperimentIO`'s `readCosmxSXE` function. 
To not confuse RNA target counts with counts stemming from either negative 
probes and system controls, the latter are moved into `altExp`s by default.
We also add additional `colData` and `metadata` slots that are expected by 
`r BiocStyle::Biocpkg("ggspavis")` (for visualization) and 
`r BiocStyle::Biocpkg("SpaceTrooper")` (for analysis). In particular, 
`SpaceTrooper`'s `updateCosmxSPE` function takes care of standardizing
the object's `colData` for use with `SpaceTrooper`'s functions.

::: {.aside}
See @sec-img-introduction for details on the three main types 
of barcodes: RNA targets, negative probes, and system controls.
:::

```{r load-data-cos, message=FALSE, warning=FALSE}
# retrieve CosMx dataset from OSF repo
id <- "CosMx1k_MouseBrain2"
pa <- OSTA.data_load(id, mol=FALSE)
dir.create(td <- tempfile())
unzip(pa, exdir=td)
cos <- readCosmxSXE(td, addTx=FALSE)

# prepare data for 'SpaceTrooper' 
cos <- updateCosmxSPE(cos, td, sampleName="CosMx")
cos <- readAndAddPolygonsToSPE(cos)
cos$in_tissue <- TRUE
cos
```

## Visualization

Imaging-based ST data relies on imaging a 2D tissue section. Hence, an advantage 
compared to scRNA-seq is that we can readily visualize a 2D map of the cells and 
directly plot cell features onto the map.
The easiest way to visualize the data is by representing the cells as centroids. 
This is useful to have a global bird's-eye-view of the tissue.

```{r plot-xy, fig.width=10, fig.height=5}
# spatial plots for datasets acquired through
# Xenium & CosMx (points = cell centroids)
plotCoords(xen, point_size=0, y_reverse = FALSE) + ggtitle(xen$sample_id) +
plotCoords(cos, point_size=0, y_reverse = FALSE) + ggtitle(cos$sample_id)
```

## General QC metrics

Similarly to scRNA-seq, quality control (QC) is an important step to remove 
low-quality cells and highlights specific biases in the data at the sample, 
field of view (FOV), or slide area level. Unlike scRNA-seq and sequencing-based 
ST, however, most imaging-based ST platforms do not measure the transcriptome, 
nor an "unbiased" sample of it, but rely on the design of gene panels. 

Depending on the platform and the application, such gene panels 
may be pan-tissue or targeting specific tissues or cell types. 
Moreover, they vary from a few hundreds (e.g., 313 in the first Xenium panel) 
to several thousands (e.g., the 5k Xenium panel or the 6k and 18k CosMx panels). 
Hence, unlike in sequencing-based platforms, we may not have enough
ribosomal or mitochondrial transcripts to use for computing QC metrics. 
However, imaging-based platforms provide a set of negative probes 
that can be used to check for aspecific RNA captures.

Finally, imaging-based platforms include, in addition to transcripts data, 
morphology information (such as area and eccentricity of segmented cells) 
and information from antibody stains (such as those targeting cell nuclei 
or surface; see above).

### scRNA-seq

We start out by computing QC metrics that are standard for scRNA-seq data,
using the `r BiocStyle::Biocpkg("scrapper")` package. Although not available 
for these data, the `altexp.proportions` argument could be used to quantify
metrics across alternative experiments, such as negative probes or blank codes, 
in addition to RNA targets, which represent the object's primary data.

::: {.aside}
See also @sec-seq-quality-control.
:::

```{r scrapper-qc}
# compute standard QC metrics for Xenium
# (i.e., total counts & unique features)
xen <- quickRnaQc.se(xen, subsets=list())
```

```{r gc-1, include=FALSE}
# garbage collection
# to reduce memory usage for building book with GitHub Actions
gc()
```

### IF staining

CosMx outputs, for instance, typically include the min/max/mean intensities 
of a set of immunofluorescence (IF) markers for, e.g., nuclei and cell surface 
(used for segmentation), as well as additional markers (experiment-dependent).
Here, we show the distribution of mean IF intensities across the CosMx sample.
[Values are arcsinh-transformed with a cofactor of 10 for visual purposes]{.aside}

```{r plot-cos-xy-dapi, message=FALSE, fig.width=12, fig.height=4}
#| code-fold: true
flu <- grepv("^Mean", names(colData(cos)))
lapply(flu, \(.) {
    cos[[.]] <- asinh(cos[[.]]/10)
    plotCoords(cos, annotate=., point_size=0, y_reverse = FALSE)
}) |>
    wrap_plots(nrow=1) &
    theme(
        legend.position="bottom", 
        legend.key.height=unit(0.4, "lines")) &
    scale_color_gradientn(colors=pals::jet())
```

```{r gc-2, include=FALSE}
# garbage collection to reduce memory usage 
# for building book with GitHub Actions
gc()
```

### Morphology

`r BiocStyle::Biocpkg("SpaceTrooper")`'s `spatialPerCellQC` function can be 
used to compute a set of metrics that are particular to imaging-based ST data; 
examples listed below. 
In addition, the function incorporates an internal call to `quickRnaQc.se`
from the `r BiocStyle::Biocpkg("scrapper")` package to compute some standard 
scRNA-seq QC metrics on an `altExp`; here, we specify using negatives probes.

::: {.aside}
Note that metrics also include each cell's distance from horizontal/vertical
and nearest FOV borders: `dist_border_x/y` and `dist_border`, respectively;
we will get back to this in @sec-img-quality-control-cos.
:::

```{r SpaceTrooper-qc, message=FALSE, warning=FALSE, echo=2:4}
old <- names(colData(cos))
# compute 'SpaceTrooper' QC metrics
cos <- spatialPerCellQC(cos, use_altexps="NegPrb")
# the following cell metadata are newly available:
setdiff(names(colData(cos)), old)
```

```{r gc-3, include=FALSE}
# garbage collection
# to reduce memory usage for building book with GitHub Actions
gc()
```

In particular, the following metrics have been added to the `colData`:

- `Area_um`: area in microns
- `SignalDensity`: count density = number of transcripts per area
- `log2AspectRatio`: log2 of the aspect ratio (x-/y-axis) of each cell
- `ctrl_total_ratio`: proportion of negative probe counts over total counts


### Counts \& area

Imaging-based ST data relies on imaging a 2D tissue section. Cells thus represent 
a slice through a 3D system, and we expect a tight relationship between a cell's 
counts and its area. When we inspect total counts and cell areas as separate 
(univariate) distributions, both appear approximately log-normal distributed.

::: {.aside}
Note that there are many more counts for the CosMx data, mostly because
there are `r nrow(cos)` features but only `r nrow(xen)` for Xenium.
:::

```{r plot-qc-1d, fig.width=10, fig.height=3, warning=FALSE}
#| code-fold: true
.p <- \(df, xs) {
    fd <- pivot_longer(df, all_of(xs))
    mu <- summarise_at(group_by(fd, name), "value", median)
    ggplot(fd, aes(value)) + facet_grid(~ name) + 
        geom_histogram(bins=50, linewidth=0.1, fill="gray") + 
        geom_vline(data=mu, aes(xintercept=value), col="blue") + 
        geom_text(
            hjust=-0.1, size=3, col="blue", 
            data=mu, aes(value, 0, label=round(value))) + 
        scale_x_continuous(NULL, trans="log10") + ylab("# cells") + 
        theme_minimal() + theme(panel.grid.minor=element_blank())
}
df_cos <- data.frame(colData(cos), spatialCoords(cos))
df_xen <- data.frame(colData(xen), spatialCoords(xen))
p2 <- .p(df_cos, c("Area_um", "total"))
p1 <- .p(df_xen, c("cell_area", "total_counts"))
p1 + ggtitle("CosMx") | p2 + ggtitle("Xenium")
```

Similarly, we can visualize the distribution of these metrics in the tissue 
by plotting cell centroids colored by total counts and cell area, respectively.

```{r plot-qc-xy, message=FALSE, fig.width=12, fig.height=4}
#| code-fold: true
lapply(c("cell_area", "total_counts"), \(.) {
    # crop very low/high values for clearer visualization
    val <- xen[[.]]
    qs <- quantile(val, c(0.01, 0.99))
    val <- ifelse(val < qs[1], qs[1], ifelse(val > qs[2], qs[2], val))
    xen[[.]] <- val
    plotCoords(xen, annotate=., point_size=0, y_reverse = FALSE) +
        theme(legend.key.width=unit(0.5, "lines")) +
        scale_color_gradientn(colors=unname(pals::jet()), trans="log10")
}) |> wrap_plots(nrow=1)
```

In a bivariate setting, we observe a positive relationship between counts and area (as expected). 
However, this dependency is not always linear and may exhibit a bimodal distribution.

```{r plot-qc-2d-one, message=FALSE, warning=FALSE, fig.width=6, fig.height=3}
#| code-fold: true
.p <- \(df, x, y) {
    ggplot(df, aes(.data[[x]], .data[[y]])) + 
        geom_point(shape=16, size=0, alpha=0.1) + 
        geom_smooth(method="lm", col="blue") +
        scale_x_log10() + scale_y_log10() +
        theme_minimal() + theme(
            aspect.ratio=1, 
            panel.grid.minor=element_blank())
}
.p(df_cos, "Area_um", "total") + ggtitle("CosMx") +
.p(df_xen, "cell_area", "total_counts") + ggtitle("Xenium")
```

In some cases, a peculiar count-area distribution might be indicative of 
segmentation problems. Undersegmentation (i.e., when groups of cells are 
merged together) might lead to fewer counts than expected for a given 
area (e.g., when they are far apart and empty space is being segmented).
[An useful metric to explore this issue is the `log2SignalDensity`, 
which estimates the transcript density for each cell.]{.aside}

## CosMx-specific QC {#sec-img-quality-control-cos}

### Fields of view (FOVs)

Imaging-based ST data from CosMx are acquired through iterative imaging 
of predefined rectangular regions, so-called fields of view (FOV).

Unlike Xenium and MERSCOPE, standard CosMx pipelines do not stitch FOVs, but 
rather treat each FOV independently. The `colData` of the `SpatialExperiment` 
object resulting from a CosMx data import will include the FOV of each cell, 
but it is sometimes tricky to understand (or remember) the relative position 
of each FOV within the slide. 

`r BiocStyle::Biocpkg("SpaceTrooper")`'s `plotZoomFovsMap` function provides an 
easy way to map FOVs within the slide, allowing to zoom in on a subset of FOVs, 
while visualizing the global structure of the tissue.

```{r plot-cos-fov, warning=FALSE, fig.width=10, fig.height=5}
plotZoomFovsMap(cos, 
    mapPointCol="green",
    colourBy="Mean.DAPI",
    fovs=c(98:100, 113:115, 122:124))
```

Assuming FOV placement aims to capture biologically interesting regions -- 
regions of interest (ROIs) -- we would expect every FOV to capture a decent 
number of cells. Issues in image registration or IF staining, however, may 
affect spot calling and cell segmentation, and can yield FOVs with virtually 
no cells. We thus advice inspecting the number of cells across FOVs:

```{r cos-plot-ns, fig.height=2, out.width="100%"}
#| code-fold: true
ns <- as.data.frame(
    table(fov=cos$fov), 
    responseName="n_cells")
ggplot(ns, aes(fov, n_cells)) + 
    scale_x_discrete(breaks=c(1, seq(5, max(cos$fov), 5))) +
    scale_y_continuous(limits=c(0, 1e3), n.breaks=4) +
    labs(x="field of view (FOV)", y="# cells") +
    geom_col(fill="grey", alpha=2/3) +
    coord_cartesian(expand=FALSE) +
    theme_bw()
```

It is generally also worth checking that basic QC metrics do not vary across FOVs.
Of course, such effects will also be driven by differences in subpopulation composition across FOVs.
However, gross differences in detection efficacy, IF staining, spot calling, segmentation, etc. often still manifest in FOV-level shifts in IF stains and/or RNA target, negative probe, system control counts.
A simple spot-check is the following type of visualization, but more thorough investigation is advisable 'in the wild', especially so in challenging types of tissue:

```{r cos-plot-qc, warning=FALSE, fig.width=10, fig.height=5}
#| code-fold: true
df <- data.frame(colData(cos))
ys <- c("detected", "total", "Area_um", "log2SignalDensity")
ys <- c(grepv("Mean", names(df)), ys)
fd <- pivot_longer(df, ys) |>
    mutate(value=case_when(
        grepl("Mean", name) ~ asinh(value),
        grepl("total", name) ~ log10(value),
        grepl("^Area", name) ~ log10(value),
        TRUE ~ value), fov=factor(fov))
ggplot(fd, aes(fov, value, fill=fov)) + 
    facet_wrap(~name, nrow=3, scales="free_y") +
    geom_boxplot(linewidth=0.2, outlier.size=0.2) +
    scale_x_discrete(breaks=c(1, seq(10, max(cos$fov), 10))) +
    theme_minimal() + theme(legend.position="none") +
    labs(x="field of view (FOV)", y=NULL)
```

In our example, DAPI stains, RNA target counts (`total`), and the number of
uniquely detected RNA targets (`detected`) are fairly constant across FOVs; 
systematic differences seem to reflect morphological differences across FOVs.
Taken together, we cannot deem any FOVs as being obviously problematic here.

### Border effects

Particularly in CosMx, lack of FOV stitching during cell segmentation can result 
in fractured and possibly duplicated cells. To investigate potential artefacts 
related to this, we can use the distance to FOV borders calculated previously.

We can *roughly* estimate cell radii as $A=\pi r^2$ $\leftrightarrow r=\sqrt{A/\pi}$, 
and use this as a threshold on FOV border distances; i.e., we flag a cell when 
its centroid is closer to any border than the radius of an average circular cell.

```{r cos-est-rad}
(r <- sqrt(mean(cos$Area)/pi))
```

Now, let us visualize total RNA counts against distance to FOV borders, 
setting the above estimate for radius `r` as a threshold on the latter; 
we also include rolling means to better highlight global trends:

```{r cos-plot-ds, message=FALSE, warning=FALSE, fig.width=9, fig.height=4}
#| code-fold: true
df <- data.frame(colData(cos))
.p <- \(df, x, y) {
    ggplot(df, aes(.data[[x]], .data[[y]])) +
        geom_point(color="grey", size=0) +
        geom_smooth() + theme_minimal() +
        geom_vline(xintercept=r, col="red") +
        geom_vline(xintercept=50, linetype=2)
}
p1 <- .p(df, "dist_border", "total") + xlim(0, 500) + ylim(0, 2000)
p2 <- .p(df, "dist_border", "Area_um") + xlim(0, 500) + ylim(0, 200)
p3 <- .p(df, "dist_border_x", "log2AspectRatio") + xlim(0, 500)
p4 <- .p(df, "dist_border_y", "log2AspectRatio") + xlim(0, 500)
(p1 + p2) / (p3 + p4) 
```

As expected, we observe a decline in counts near FOV borders, with about 
`r round(100*mean(df$dist_border < r), 1)`% of cells falling below $r$ (red line) and 
`r round(100*mean(df$dist_border < 50), 1)`% of cells falling below 50 pixels (dashed line).

These cells exhibit lower total counts and smaller segmented areas, 
as well as a distorsion in their aspect ratio (thin cells close to 
the y border and broad cells close to the x boder).

To mitigate potential artefacts in downstream analyses, we may choose to filter 
out these cells, or otherwise flag them as potentially problematic events to be 
kept a cautionary eye on. 

```{r cos-plot-ex, warning=FALSE, fig.width=5, fig.height=5}
#| code-fold: true
#| layout-ncol: 2
# highlight border cells across 
# the tissue (using cell centroids)
cos$ex <- df$dist_border < r
plotCentroids(cos, colourBy="ex", sampleId=NULL) +
    scale_color_manual(values=c("lavender", "blue")) +
    guides(col=guide_legend(override.aes=list(size=2))) +
    theme_void() + theme(legend.key.size=unit(0, "pt"))
    
# zoom in on subset of FOVs
# visualizing cell boundaries
fs <- c(98:100, 113:115, 122:124)
plotPolygons(
    cos[, cos$fov %in% fs], 
    colourBy="ex", sampleId=NULL,
    borderCol="black", palette=c("lavender", "blue")) +
    theme(legend.position="none")
```

## Identifying low-quality cells

We can now use the QC metrics calculated above to flag 
(and potentially remove) low-quality cells.

### Individual metrics

First, we can consider metrics separately looking for outliers in their distribution.
Depending on the metric, we are interested in low-end or high-end outliers (or both). 
For instance, we may want to exclude too small/large cells as potential artifacts.

`r BiocStyle::Biocpkg("SpaceTrooper")` provides a function to compute outliers 
of the distribution of any QC metric present in the `colData` of the object. 
We can then visualize the distribution and outlier threshold with a histogram.

::: {.aside}
Note that thresholds could also be obtained using `scuttle`'s `isOutlier` 
function on the corresponding metric(s); see @sec-img-quality-control-fil.
:::

```{r SpaceTrooper-ol-cos, fig.height=2.5, fig.width=10}
cos <- computeSpatialOutlier(cos, computeBy="total", method="mc")
cos <- computeSpatialOutlier(cos, computeBy="Area_um", method="mc")
plotMetricHist(cos, metric="total", useFences="total_outlier_mc") +
plotMetricHist(cos, metric="Area_um", useFences="Area_um_outlier_mc")
```

In the CosMx data, we identify `r sum(cos$total_outlier_mc != "NO")` outliers 
for total counts and `r sum(cos$Area_um_outlier_mc != "NO")` outliers for area.

### Combined score

Finally, we can compute an aggregated quality score, which combines a cell's 
transcript density, aspect ratio and (for CosMx) distance from FOV borders.

Specifically, `r BiocStyle::Biocpkg("SpaceTrooper")`'s `computeQCScore` function will compute a combined quality score based on count density (`log2SignalDensity`), cell aspect ratio (`log2AspectRatio`), only for cells closer than a specified distance from FOV borders, and control-to-total count ratio (`log2Ctrl_total_ratio`). The count density is computed as the ratio between the per cell total gene counts divided by the cell area; the aspect ratio represents FOV border effect typical of CosMx datasets; the control-to-total ratio is the ratio between counts for control probes and total counts per cell and represents aspecific signal.

::: {.aside}
The quality score is computed using a ridge logistic regression model, which predicts a score between 0 and 1, where higher scores indicate better quality cells, using the following model forumla: `~(log2SignalDensity + Area_um + log2Ctrl_total_ratio + I(abs(log2AspectRatio) * as.numeric(dist_border<50)))^2`, where the `^2` notation indicates all pairwise interactions among the included variables. See `?computeQCScore` for details.
:::

Here, `computeQCScore()` will add a (numeric) `QC_score`, 
and `computeQCScoreFlags` will add a (logical) `low_qcscore` slot
to the input SPE's `colData`; the latter flags low-quality cells.

```{r SpaceTrooper-calc-score, message=FALSE}
# compute quality scores & flag cell's that
# fall below specified percentile threshold
cos <- computeQCScore(cos)
cos <- computeQCScoreFlags(cos, 
    qsThreshold=0.5)

# the following slots have now been 
# added to the object's cell metadata
nms <- names(colData(cos))
sco <- grepv("score$", nms)
head(colData(cos)[sco])
```

```{r SpaceTrooper-plot-score, message=FALSE, fig.width=10, fig.height=5}
#| code-fold: true
# visualize quality scores & highlight flagged cells
thm <- list(theme_void(), ggtitle(""), coord_equal())
plotCentroids(cos, colourBy="QC_score") + thm +
plotCentroids(cos, colourBy="low_qcscore") + thm +
    guides(col=guide_legend(override.aes=list(size=2))) +
    scale_color_manual(values=c("lavender", "blue")) +
    theme(legend.key.size=unit(0, "pt")) 
```

We can see how the QC score flags several cells close to the tissue border as well as cells close to the FOVs borders. 

## Filtering {#sec-img-quality-control-fil}

For the Xenium dataset, we will employ rather stringent filtering criteria to:

- exclude cells with too few counts per area (thresholding on MADs)
- exclude cells with *any* negative probe or system control counts

```{r xen-ol}
# estimate threshold on counts per area
# (3 MADs from the median, using log2 scale)
nc <- xen$total_counts / xen$cell_area
ol <- isOutlier(nc, log=TRUE, type="lower", nmads=3)
(th <- attr(ol, "threshold")[1])
```

```{r xen-plot-ol, fig.width=4, fig.height=3}
#| code-fold: true
# visualize histrogram of counts 
# per area, including threshold
par(mar=c(4, 4, 0, 0))
hist(log(nc), 
    n=50, col="lavender",
    main=NULL, ylab="# cells", 
    xlab="log(total_counts / cell_area)")
abline(v=log(th), col="blue")
```

```{r xen-ex}
# fraction of low-quality Xenium cells
mean(ex <- ol | 
    xen$control_probe_counts > 0 | 
    xen$control_codeword_counts > 0)
```

For the CosMx dataset, we might filter based on `SpaceTrooper`s quality score,
which would exclude about `r round(100*mean(cos$low_qcscore))`\% of cells.

```{r cos-ex}
# fraction of low-quality CosMx cells
mean(cos$low_qcscore)
```

## Conclusion

In a stringent analysis, we may want to remove low-quality cells, but this has 
to be done with caution: **before excluding any cells, it is recommendable to 
inspect where these fall in the tissue**. We would expect low-quality cells to 
be distributed randomly, and otherwise to accumulate at the tissue or FOV borders or 
in regions were tissue was, for example, smudged, detached, necrotic etc.

In general, **spatially organized clusters of excluded cells might indicate a 
bias towards excluding specific types of cells** (e.g., some immune cells tend to 
be smaller or might otherwise contain fewer transcripts due to the panel design).
Whenever possible, we advice readers to cross-check visualizations like the 
following with a corresponding H&E stain of the tissue, considering possible 
experimental faults as well as their prior knowledge on tissue pathology.

```{r plot-ex, echo=FALSE, fig.width=11, fig.height=5}
# 'ggspavis' complains about duplicated entries
i <- setdiff(
    names(cd <- colData(cos)), 
    spatialCoordsNames(cos))
colData(cos) <- cd[i] 
# add for visualization
xen$ex <- ex; cos$ex <- cos$low_qcscore
plotCoords(xen, annotate="ex", point_size=0, y_reverse = FALSE) + ggtitle("Xenium") +
plotCoords(cos, annotate="ex", point_size=0, y_reverse = FALSE) + ggtitle("CosMx") +
    plot_layout(guides="collect") &
    theme(legend.key.size=unit(0, "pt")) &
    scale_color_manual(values=c("lavender", "blue")) &
    guides(col=guide_legend(override.aes=list(size=2)))
xen$ex <- cos$ex <- NULL # drop for saving
```

## Appendix

::: {.callout-tip icon="false" title="Further reading"}

- Built around the `SpatialData` framework [@Marconato2025-SpatialData],
`r BiocStyle::Githubpkg("LazDaria/SegTraQ")` is a Python-based toolkit
for **assessing quality** of cell segmentation and transcript assignment.

- @Ishaque2026 provide a perspective on **segmentation challenges**
and how they impact downstream analysis, and discuss potential
future directions for benchmarking segmentation algorithms
(see also @sec-img-segmentation).

- Bruker's [CosMx Analysis Scratch Space](
https://nanostring-biostats.github.io/CosMx-Analysis-Scratch-Space/about.html) 
provides an exploratory (open-source) **resource for CosMx** data analysis, which
includes an FOV-level QC approach that relies on comparing proximal 
FOVs in order to identify outliers in terms of RNA detection
(see [here](https://nanostring-biostats.github.io/CosMx-Analysis-Scratch-Space/posts/fov-qc));
and, a theoretical piece on *background* signal -- i.e., non-RNA target binding events
(see [here](https://nanostring-biostats.github.io/CosMx-Analysis-Scratch-Space/posts/background)).

:::

### Save data

```{r save-data}
# save post-QC Xenium data
# for consecutive chapters
saveRDS(xen[, !ex], "img-spe_qc.rds")
```

### References {.unnumbered}

{{< include _stop.qmd >}}
