# Clustering & annotation {#sec-ind-clustering}

{{< include _start.qmd >}}

## Preamble

### Introduction

In spatial transcriptomics data, we can apply clustering algorithms to identify spatial domains, which represent spatially defined regions consisting of relatively consistent gene expression profiles. 
For example, spatial domains may consist of regions containing cells from a single cell type or a consistent mixture of cell types.

Several alternative approaches exist for identifying spatial domains. One strategy involves applying standard clustering algorithms from single-cell workflows without incorporating spatial information. 
Alternatively, spatial information can be integrated at various stages of the workflow. 
For example, spatial data can inform preprocessing steps, such as selecting spatially variable genes (SVGs), followed by the application of either non-spatial or spatial clustering algorithms. 
Another widely used strategy involves generating a latent space that incorporates spatial coordinates, often leveraging Bayesian framework or neural networks.

As advancements in spatial transcriptomics techniques enhance tissue resolution, computational tradeoffs between various approaches and parameters become increasingly relevant.

It is also important to keep in mind that when we use clustering to define cell types and/or states, these can be defined at various resolutions (or even on a continuum). 
The optimal number of clusters depends on the biological context -- in particular, there is no "true" number of clusters, since this depends on the biological context (e.g. if we are comparing major cell populations vs. comparing rare subtypes), so the choice of the optimal number of clusters or resolutions requires some judgment and biological interpretation.

Once we have identified spatial domains, these can then be further investigated in additional downstream analyses.

Spatial domain identification is one component of a broader analysis workflow.
Other related tasks include classifying cellular neighborhoods or niches based on local cell-type composition (see @sec-img-neighborhood-analysis), identifying spatially variable genes (see @sec-ind-feature-selection-testing-svgs), estimating cell-type mixtures in spots (see @sec-seq-deconvolution), and modeling spatially constrained cell-cell communication (see @sec-img-cell-cell-communication).
In this chapter, we focus on clustering cells or spots into spatial domains, while pointing readers to complementary chapters for other related analyses.

### Dependencies

```{r deps, message=FALSE, warning=FALSE}
library(Banksy)
library(BayesSpace)
library(basilisk)
library(ggspavis)
library(OSTA.data)
library(patchwork)
library(pheatmap)
library(reticulate)
library(scrapper)
library(SpatialExperiment)
# set seed for random number generation
# in order to make results reproducible
set.seed(2025)
# load processed Xenium datasets
spe <- readRDS("img-spe_cl.rds")
```

## Clustering

### Non-spatial

As a baseline, we will perform a standard graph-based clustering approach developed for scRNA-seq data, using molecular features (gene expression) only.
Specifically, we construct a shared nearest neighbor (SNN) graph based on non-spatial PCs (`reducedDim` slot `"PCA_tx"`), and use the Leiden algorithm for community detection [@Traag2019-Louvain-to-Leiden].
[The `resolution` parameter used here was selected in order to obtain a similar number of clusters annotated by the authors.]{.aside}

```{r clu-Leiden}
# build cellular shared nearest-neighbor (SNN) graph;
# cluster using Leiden community detection algorithm
spe <- clusterGraph.se(spe,
    reddim.type="PCA_tx", 
    output.name="Leiden",
    method="leiden", 
    resolution=1.2, 
    num.neighbors=20,
    more.build.args=list(weight.scheme="jaccard"))
table(spe$Leiden)
```

### Spatially-aware

Several methods to identify spatial domains in ST data have recently been developed, which each have various methodological and computational tradeoffs. 
A few example include `r BiocStyle::Biocpkg("BayesSpace")` [@Zhao2021-BayesSpace], `r BiocStyle::Biocpkg("Banksy")` [@Singhal2024-BANKSY] and `r BiocStyle::CRANpkg("PRECAST")` [@Liu2023-PRECAST] in R, as well as `r BiocStyle::Githubpkg("CellCharter")` [@Varrone2024-CellCharter] and `r BiocStyle::Githubpkg("SpaceFlow")` [@Ren2022-SpaceFlow] in Python.

Here, we demonstrate tools representative of different methodologies, namely, (probabilistic) `BayesSpace` for seq-based Visium, as well as (neighborhood-based) `Banksy` and (encoder-based) `CellCharter` for img-based Xenium data.

#### Probabilistic

Underlying `r BiocStyle::Biocpkg("BayesSpace")` [@Zhao2021-BayesSpace] is a Bayesian model that encourages neighboring spots to belong to the same cluster, and utilizes Markov Chain Monte Carlo (MCMC) for estimating model parameters.
[Default parameters are `nrep=50,000` MCMC iterations and a burn-in period of `burn.in=1,000`; the authors recommend running with at least 10,000 iterations. Here, we use fewer iterations for the sake of runtime, which will arguably compromise performance. Note that, since MCMC is stochastic, a seed for random number generation should be set in order to make results reproducible.]{.aside}

```{r BayesSpace, message=FALSE}
# load processed Visium dataset
vis <- readRDS("seq-spe_cl.rds")

# prepare data for 'BayesSpace'
# skipping PCA (already computed)
.vis <- spatialPreprocess(vis, skip.PCA=TRUE)

# perform spatial clustering with 'BayesSpace'
# using 'd=20' PCs and targeting 'q=10' clusters
.vis <- spatialCluster(.vis, q=10, d=20, nrep=1e3, burn.in=100) 
table(vis$BayesSpace <- factor(.vis$spatial.cluster))
```

#### Neighborhood-based

`r BiocStyle::Biocpkg("BANKSY")` [@Singhal2024-BANKSY] embeds cells in a product space of their own and their local neighborhood's (average) transcriptome, representing cell state and microenvironment.
We have already computed `Banksy` PCs in @sec-ind-dimensionality-reduction (`reducedDim` slot `PCA_sp`) so that here, we perform SNN graph-based Leiden clustering on said PCs in order to obtain spatially-aware cluster assignments:
[Note that we are using a lower `resolution` here, as `Banksy` PCs tend to yield more clusters on the data at hand.]{.aside}

```{r clu-Banksy, message=FALSE}
# perform SNN graph-based clustering on 'Banksy' PCs using
# similar parameters as for 'non-spatial' clustering above
spe <- clusterGraph.se(spe,
    reddim.type="PCA_sp", 
    output.name="Banksy",
    method="leiden", 
    resolution=0.8, 
    num.neighbors=20,
    more.build.args=list(weight.scheme="jaccard"))
table(spe$Banksy)
```

#### Encoder-based

Another category of spatially-aware clustering method uses encoder architectures to generate a latent embedding incorporating both spatial and transcriptomics information, which is then clustered using standard algorithms. 
One typical example is `r BiocStyle::Githubpkg("CSOgroup/cellcharter")` [@Varrone2024-CellCharter], which uses a variational auto-encoder to generate a latent feature space and aggregate the features of each spot across its neighbors to preserve their spatial context. 

Recent machine-learning approaches for spatial domain identification often construct a graph whose nodes represent spots or cells and whose edges encode spatial proximity, transcriptional similarity, histological similarity, or combinations thereof.
Graph neural networks, graph attention autoencoders, variational autoencoders, contrastive learning, and foundation model embeddings can then be used to learn spatially informed representations before clustering.
These methods can be powerful when tissue structure depends jointly on expression, local neighborhood composition, and histology, but they also introduce additional dependencies, hyperparameters, hardware requirements, and reproducibility considerations.

::: {.callout-note title="Other methods" collapse="true"}
Several other published methods have also adopted encoder architectures in different ways to perform spatially-aware clustering. 
`r BiocStyle::Githubpkg("zhanglabtools/STAGATE")` [@Dong2022-STAGATE] uses a graph attention autoencoder to integrate spatial and transcriptional data by modeling spatial neighborhoods as graphs, whereas `r BiocStyle::Githubpkg("Tang-Lab-super/PROST")` [@Liang2024-PROST] employs a self-attention transformer-based encoder to learn spatial patterns by jointly modeling gene expression and spatial coordinates. 
Graph-based encoder methods, such as `STAGATE`, may be particularly well-suited for imaging-based ST data, as their ability to model spatial neighborhoods as graphs aligns naturally with the continuous and fine-grained spatial organization of tissues captured through imaging.
Other examples include `r BiocStyle::Githubpkg("jianhuupenn/SpaGCN")` [@Hu2021-SpaGCN], `r BiocStyle::Githubpkg("JiangBioLab/DeepST")` [@Xu2022-DeepST], `r BiocStyle::Githubpkg("JinmiaoChenLab/GraphST")` [@Long2023-GraphST], `r BiocStyle::Githubpkg("hongleir/SpaceFlow")` [@Ren2022-SpaceFlow], `r BiocStyle::Githubpkg("Lotfollahi-lab/nichecompass")` [@Birk2025-NicheCompass], and `r BiocStyle::Githubpkg("MICS-Lab/novae")` [@Blampey2025-Novae].
We demonstrate `CellCharter` here as one representative encoder-based workflow, and recommend benchmark and review papers below for method selection.
:::

```{r env, warning=FALSE, eval=FALSE}
# initialize 'basilisk' environment
env <- BasiliskEnvironment(
  envname="CellCharter", pkgname="OSTA",
  channels=c("conda-forge", "pytorch"),
  packages=c(
    "python=3.10.15",
    "pytorch=1.12.1",
    "torchvision=0.13.1",
    "torchaudio=0.12.1"),
  pip=c(
    "scvi-tools==0.20.3",
    "cellcharter==0.2.0",
    "anndata==0.10.9",
    "scanpy==1.10.4"))
# activate underlying conda environment
use_condaenv(obtainEnvironmentPath(env))
counts <- r_to_py(t(counts(spe)))
coords <- r_to_py(spatialCoords(spe))
```

```{python CellCharter, eval=FALSE}
import cellcharter as cc
import anndata as ad
import squidpy as sq
import pandas as pd
import numpy as np
import random
import scvi

adata = ad.AnnData(X = r.counts,
                   obsm = {"spatial":r.coords},
                   layers = {"counts": r.counts})

seed = 2025
random.seed(seed)
scvi.settings.seed = seed

# variational autoencoder for feature extraction
scvi.model.SCVI.setup_anndata(adata, layer="counts")
model = scvi.model.SCVI(adata, n_hidden=64)
model.train(early_stopping=True,
    # the parameters below aim to reduce runtime; 
    # in reality, it'd be better to use defaults
    max_epochs=70, batch_size=512, train_size=0.5, validation_size=0.2)

adata.obsm["X_scVI"] = model.get_latent_representation(adata).astype(np.float32)

# Getting neighborhood aggregation
sq.gr.spatial_neighbors(adata, coord_type="generic", delaunay=True, spatial_key="spatial", percentile=99)
cc.gr.aggregate_neighbors(adata, n_layers=3, use_rep="X_scVI", out_key="X_cellcharter")

# clustering by scanning a range of data number
mod = cc.tl.Cluster(n_clusters=14, random_state=seed)
mod.fit(adata, use_rep="X_cellcharter")

label_df = pd.DataFrame({"label": mod.predict(adata, use_rep="X_cellcharter")}) 
label_df[["label"]].value_counts()
```

```{r py2r, eval=FALSE}
# pull 'CellCharter' assignments from Python into R
spe$CellCharter <- unlist(py$label_df)
```

## Downstream

### Visualization

Let's visualize the assignment obtains from non-spatial and spatially-aware clustering:

```{r plot-xy-k-joint, message=FALSE, fig.width=11, fig.height=6}
#| code-fold: true
ks <- c("Label", "Leiden", "Banksy")
lapply(ks, \(.) {
    plt <- plotCoords(spe, annotate=.)
    plt$layers[[1]]$aes_params$stroke <- 0
    plt$layers[[1]]$aes_params$size <- 0.2
    plt
}) |>
    wrap_plots(nrow=2) &
    scale_color_manual(values=unname(pals::trubetskoy())) &
    theme(legend.key.size=unit(0, "lines"), legend.justification="left")
```

Especially in large tissues, and when there are many subpopulations, the above type of plot makes it difficult to spot rare subpopulations, and might cause cells to overlap in regions with high cellular density.
This can be misleading, as we will tend to see only highly abundant subpopulations, or the cells plotted last and on top (i.e., later columns in the object).

To better distinguish between different subpopulations, we can instead generate separate spatial plots with one subpopulation highlighted at a time:

```{r plot-xy-k-split, fig.width=9, fig.height=6}
#| code-fold: true
# plot selected clusters in order of frequency,
# highlighting cells assigned to cluster 'k'
lapply(tail(names(sort(table(spe$Banksy))), 12), \(k) {
    spe$foo <- spe$Banksy == k
    spe <- spe[, order(spe$foo)]
    plt <- plotCoords(spe, annotate="foo")
    plt$layers[[1]]$aes_params$stroke <- 0
    plt$layers[[1]]$aes_params$size <- 0.2
    plt + ggtitle(k)
}) |>
    wrap_plots(nrow=3) &
    scale_color_manual(values=c("lavender", "purple")) &
    theme(plot.title=element_text(hjust=0.5), legend.position="none")
```

::: {.callout-note title="PC regression" collapse="true"}

### PC regression

For any single-cell analysis where downstream tasks rely on PCs, it is useful to perform linear regression of (continuous or categorical) covariates of interest onto PCs. 
This quantifies the variance explained by the covariate and can help assess the extend of unwanted variation (due to, e.g., cell area) as opposed to subpopulations driving transcriptional differences. 
Here, we regress total counts, cell area, and cluster assignments from different methods against PCs:

```{r pcr}
pcs <- reducedDim(spe, "PCA_tx")
ids <- c("total_counts", "cell_area", ks)
pcr <- lapply(ids, \(id) {
    fit <- summary(lm(pcs ~ spe[[id]]))
    r2 <- sapply(fit, \(.) .$adj.r.squared)
    data.frame(id, pc=seq_along(r2), r2)
}) |> do.call(what=rbind)
```

Here, `Leiden` (transcription-only) and `Banksy` (spatially-aware) clusterings perform similar in terms of capturing (spatially unaware) PCs; as to be expected, `Leiden` clusters do slightly better.
Had we set a higher value for $\lambda$ when running `Banksy`, results would diverge more; vice versa, using spatial PCs for regression would have `Leiden` clusters perform worse.
Also, note that we would expect results to converge for $\lambda=0$.

```{r plt-pcr, warning=FALSE, fig.width=5, fig.height=2.5}
#| code-fold: true
pcr$id <- factor(pcr$id, ids)
pal <- c("red", "magenta", "gold", "cyan", "blue", "black")
ggplot(pcr, aes(pc, r2, col=id)) +
    geom_line(show.legend=FALSE) + geom_point() +
    scale_color_manual("predictor", values=pal) +
    scale_x_continuous(breaks=c(1, seq(5, 20, 5))) +
    scale_y_continuous(limits=c(0, 1), breaks=seq(0, 1, 0.2)) +
    labs(x="principal component", y="coeff. of determination") +
    guides(col=guide_legend(override.aes=list(size=2))) +
    coord_cartesian(xlim=c(1, 20)) +
    theme_minimal() + theme(
        panel.grid.minor=element_blank(),
        legend.key.size=unit(0, "lines"))
```

:::

### DGE analysis

To help characterize (unsupervised) clusters, we want to identify 'markers' for each subpopulation, i.e., features whose expression is positively or negatively restricted to (a) specific subpopulation(s).
For details on identifying genes that are differentially expressed (DE) between groups of cells, we refer readers to [OSCA](https://bioconductor.org/books/3.19/OSCA.basic/marker-detection.html); a standard approach is given below, visualizing the average expression of exemplary markers across clusters:

```{r mgs}
# differential gene expression analysis
de <- scoreMarkers.se(spe, groups=spe$Leiden)

# select for a few markers per cluster
gs <- lapply(de, \(df) head(rownames(df), 7))
length(gs <- unique(unlist(gs)))
```

```{r plt-mgs-hm, fig.width=10, fig.height=3.5}
# average expression by clusters
pbs <- aggregateAcrossCells.se(spe[gs, ], spe$Leiden, assay.type="logcounts")
mtx <- sweep(assay(pbs, "sums"), 2, pbs$counts, `/`)

# visualize averages z-scaled across clusters
pheatmap(mat=t(mtx), scale="column", breaks=seq(-2, 2, length=101))
```

## Annotation

Following clustering and marker gene identification, the next step is the 
biological interpretation of identified subpopulations, typically by 
assigning them cell type or -- in spatial context -- tissue domain labels. 
Depending on data characteristics (e.g., spatial resolution, molecular coverage), 
and the availability of a suitable single-cell reference dataset, annotation 
can be performed using a range of approaches, ranging from manual to automated 
marker-based annotation, reference mapping, deep learning, or foundation models. 
Several common annotation strategies are outlined below.

- **Deconvolution**: 
For spot-based ST (e.g., Visium) where spots contain multiple cells, it is
typically required to estimate the proportions of different cell types within 
each spot through reference-based deconvolution (see @sec-seq-deconvolution).

- **Manual**: 
Clusters can be labeled based on the expression of canonical marker genes and 
results from DE analysis (as shown above). While highly accurate when performed 
by experts, this process is labor-intensive and arguably subjective.

- **Marker-based**: 
To reduce manual effort and improve reproducibility, clusters can be scored 
against predefined marker gene signatures (see @sec-ind-feature-set-signatures)
using tools such as `r BiocStyle::Biocpkg("CellAssign")` [@Zhang2019-CellAssign]
and `r BiocStyle::CRANpkg("scType")` [@Ianevski2022-ScType].

- **Reference-based**:
For imaging-based ST data in particular (e.g., Xenium, MERFISH), labels from a 
high-quality, annotated scRNA-seq reference atlas can be "projected" onto the 
spatial data. Popular R-based tools include `r BiocStyle::Biocpkg("SingleR")` 
[@Aran2019-SingleR] (Bioconductor), and `r BiocStyle::Githubpkg("satijalab/azimuth")` 
[@Hao2021-Seurat] (implemented as part of Seurat v5+).

- **Machine & deep learning**:
These methods leverage either high-capacity statistical models or multi-layered 
neural networks (e.g., VAEs, GNNs) to learn complex cellular representations. 
`r BiocStyle::Githubpkg("Teichlab/CellTypist")` [@DominguezConde2022-CellTypist],
for instance, uses a logistic regression framework and provides a variety of 
pre-trained models; `r BiocStyle::Githubpkg("scverse/scvi-tools")`'s `scANVI` 
[@Xu2021-scANVI] employs semi-supervised deep learning for label prediction.

- **Foundation models**:
More recently, foundation models (FMs) pretrained on very many single cells have emerged. 
Such models (e.g., `scFoundation` [@Hao2024-scFoundation], `scGPT` [@Cui2024-scGPT], 
and `Geneformer` [@Theodoris2023-Geneformer]) can be used zero-shot or fine-tuned 
to classify cell types across diverse tissues and platforms with high accuracy.

### Reference datasets

Some annotation approaches rely on high-quality single-cell reference datasets. 
In this context, so-called *atlas* efforts aim to generate expert-curated, 
standardized resources that serve as biological anchors across diverse species, 
tissues, developmental stages, and disease states. These resources typically 
provide large collections of molecular profiles, either through dedicated 
publications or via centralized data portals that harmonize datasets from 
multiple studies. Below, we highlight several widely known resources, 
including ongoing efforts towards spatial and multi-modal ones.

- **CELLxGENE** (Chan Zuckerberg Initiative) is a data portal and explorer 
that provides standardized access to thousands of datasets [@Megill2021-cellxgene]. 
These can be interactively queried and filtered for downstream tasks including,
e.g., reference-based annotation, benchmarking, and cross-study integration.
[The Bioconductor package `r BiocStyle::Biocpkg("cellxgenedp")` provides an 
alternative, R-based interface, allowing programmatic discovery and retrieval
of >2000 datasets from 100s of studies as H5AD files, which can be read as 
`SingleCellExperiment` objects using `r BiocStyle::Biocpkg("anndataR")` 
[@Deconinck2026-anndataR]; see @sec-bkg-python-interoperability.]{.aside}

- The **Human Cell Atlas (HCA)** is a global consortium that is mapping 
every cell type in the human body" across health and disease [@Regev2017-HCA]. 
As part of this effort, multi-organ atlases such as *Tabula Muris* [@TabulaMuris2018],
*Tabula Muris Senis* [@TabulaSenis2020] and *Tabula Sapiens* [@TabulaSapiens2022] 
provide high-resolution references for mouse and human, respectively.
[There are several R/Bioconductor `ExperimentHub` packages that provide programmatic
access to these atlases, e.g., `r BiocStyle::Biocpkg("TabulaMurisData")` and 
`r BiocStyle::Biocpkg("TabulaMurisSenisData")`. The `ExperimentHub::query()` 
command (R) or the [BiocViews browser](https://bioconductor.posit.co/packages/release/BiocViews.html#___ExperimentData) (online)
can be used to explore current datasets -- for annotation or otherwise.]{.aside}

- The **Human Protein Atlas (HPA)** integrates transcriptomics and proteomics
data to map their spatial distributions in human tissues and cells [@Thul2018-HPA].

- Still work in progress, the **Spatial Atlas of Human Anatomy (SAHA)** is 
an effort to generate a multi-modal, subcellular-resolution spatial reference 
map for millions of cells in their native 3D tissue context [@Park2025-SAHA].

### Evaluation

While automated annotation can greatly accelerate analysis, the quality of labels 
obtained through such methods depends heavily on the alignment between query and 
reference. Misalignment can arise from technical effects, biological differences 
(e.g., healthy vs. diseased), or the presence of "out-of-reference" cell types. 
Evaluating **reference suitability** and **prediction reliability** is crucial. 
The `r BiocStyle::Biocpkg("scDiagnostics")` [@Christidis2026-scDiagnostics] 
R/Bioconductor package provides a systematic framework for this task, e.g.:

- **Assessing distributional congruence** by projecting query data into 
the reference's subspace to check global alignment between datasets.

- **Marker gene validation** by quantitatively comparing the expression 
profiles of canonical markers between query and reference clusters to 
detect gene-level discrepancies.

- **Detection of annotation anomalies** by identifying cells that do not fit 
the reference; these may represent novel cell states or technical artifacts.

## Appendix

When choosing among spatial clustering methods, we recommend starting from the biological question and data modality rather than from model complexity alone.
Simpler spatially aware approaches may be preferable when interpretability, runtime, and reproducibility are priorities.
Deep learning and graph-based methods may be more appropriate when histology, multi-modal features, or complex local tissue structure are central to the analysis.
In all cases, spatial domains should be evaluated by marker genes, SVGs, histological concordance, parameter stability, and consistency with known tissue architecture where available.

::: {.callout-tip icon="false" title="Further reading"}

- The [single-cell best practices](https://www.sc-best-practices.org/spatial/domains) chapter on spatial domains divides methods into two broad categories: (i) methods that identify spatial domains using gene expression together with spatial coordinates, and (ii) methods that additionally incorporate features extracted from histological images.
Previous chapters also cover single-cell clustering and annotation using Python/scverse.

- The [OSCA](https://bioconductor.org/books/release/OSCA.basic/clustering.html) chapter on clustering provides an overview of graph-based, hierarchical, and centroid-based clustering methods for scRNA-seq data. Together with the preceding chapter on dimensionality reduction, these concepts underpin many spatial clustering approaches. Marker gene detection and cell type annotation are covered in the following couple chapters.

- @Wang2024-review-spatial-domain categorize ST clustering approaches into probability statistics-, graph neural network- and contrastive learning-based approaches, and review their advantages and limitations for clustering of data from seq- and img-based assays.

- @Sun2025-benchmark-clustering compare 22 methods across 15 datasets spanning 9 technologies and diverse tissue types. Notably, the authors also propose a "consensus-guided workflow [...] to generate consensus representations", moving beyond traditional scoring and comparison of methods based on ground truths.

- @Chen2025-benchmark-clustering compare 14 methods across ~600 datasets from 10 technologies. Several other benchmarking studies have also compared different subsets of methods across different subsets of  platforms and tissue types [@Cheng2022-benchmark-clustering; @Hu2024-benchmark-clustering; @Liu2024-network-clustering; @Xiong2025-benchmark-clustering].
 
:::

### References {.unnumbered}

{{< include _stop.qmd >}}
