# Spatial statistics {#sec-ind-spatial-statistics}

{{< include _start.qmd >}}

## Preamble

### Introduction

Spatial statistics builds around the first law of geography of @Tobler1970-urban-growth that states "everything is related to everything else, but near things are more related than distant things". This dependence of spatial observations has been studied in the field of (geo)spatial statistics. In general, spatial dependence is estimated by comparing the values at one location with the values at another location that is a given distance ('spatial lag') away [@Dale2014-book-spatial-analysis; @Baddeley2015-book-point-patterns].

The two technological streams, imaging- and sequencing-based assays, are very different in terms of data modalities: 

- In **imaging-based assays** (e.g., CosMx and Xenium) the observations of interest (e.g., mRNAs) are recorded where they occur natively. This means the locations of the observations are governed by a stochastic process and can be approximated as a **point process** or as an **irregular lattice**. 
- In **sequencing-based assays** (e.g., Visium), however, measurements are made along a defined grid, or **lattice**. This lattice is not created by a stochastic process and therefore cannot be approximated as a point process. Here, lattice data analysis methods have to be used. Notably, there are methods where cells can be segmented from very fine bins (e.g., Visium HD and IMC); these technologies can be approximated as being generated by a point process after segmentation [@Emons2025-pasta; @Baddeley2015-book-point-patterns; @Pebesma2023-book-spatial-data-science]. 

In the following vignette, the two main exploratory spatial statistics streams, point pattern analysis and lattice data analysis will be introduced.

::: {.callout-note icon="false" title="Statistical inference under spatial autocorrelation" collapse="true"}
Spatially-resolved gene expression measurements at nearby locations are typically not independent; they exhibit spatial autocorrelation. This violates independence assumptions that are assumed by many classical statistical tests (t-tests, linear regression, PCA), generally leading to underestimated uncertainty and therefore inflated false-positive rates.

As readers will notice in the many approaches proposed for spatial omics data analysis (in and outside of this book), several fundamental strategies are employed. Examples include basic methods such as Moran's $I$ (global for overall clustering, local for hotspots) to quantify directly the spatial dependence [@Moran1950-stochastic]; statistical models such as `SpatialDE` [@Svensson2018-SpatialDE], which implements correlation as a function of distance or `MIMIC` [@Gerber2026-MIMIC], which implements spatially-correlated errors according to neighbors, within standard normal or linear models, respectively; Gaussian process models capture scale-dependent spatial structure and provide uncertainty estimates that respect spatial dependence [@Weber2023-nnSVG]. 

As in all statistical inference, other themes include sufficient sample size to have statistical power (i.e., correlated observations do not represent independent observational units) and considerations for multiple testing still need to be made.
:::

### Dependencies

```{r deps, message=FALSE, warning=FALSE}
library(dplyr)
library(scran)
library(spdep)
library(tidyr)
library(ggplot2)
library(Voyager)
library(SFEData)
library(spatstat)
library(openxlsx)
library(spatialFDA)
library(BiocParallel)
library(STexampleData)
library(SpatialExperiment)
library(SpatialFeatureExperiment)
# specify whether/how to 
# perform parallelization
bp <- MulticoreParam(4)
# set seed for random number generation
# in order to make results reproducible
set.seed(77)
```

```{r, message=FALSE, warning=FALSE}
# load dataset as SPE & convert to SFE
spe <- Janesick_breastCancer_Xenium_rep1()
sfe <- toSpatialFeatureExperiment(spe)

# load the official 10X annotations 
fnm <- "https://cdn.10xgenomics.com/raw/upload/v1695234604/Xenium%20Preview%20Data/Cell_Barcode_Type_Matrices.xlsx"
labels <- read.xlsx(fnm, sheet=4)
labels$cell_id <- (labels$Barcode)

# add the cell type labels to the spe
cd <- as.data.frame(colData(sfe))
cd <- left_join(cd, 
    as.data.frame(labels), 
    by=join_by("cell_id"))
colData(sfe) <- DataFrame(cd)

# exclude 0-count cells &
# cells without annotation
sfe <- sfe[, colSums(counts(sfe)) > 0]
sfe <- sfe[, !is.na(sfe$Cluster)]

# log-library size normalization
sfe <- logNormCounts(sfe)

# basic theme for spatial plots
xy <- spatialCoords(sfe)
theme_xy <- list(
    coord_equal(expand=FALSE), 
    theme_void(), theme(
        plot.margin=margin(l=5),
        legend.key=element_blank(),
        panel.background=element_rect(fill="black")))
```

{{< include ind-spatial-statistics-la.qmd >}}

{{< include ind-spatial-statistics-pp.qmd >}}

## Appendix

::: {.callout-tip icon="false" title="Further reading"}

- This chapter has been adapted from the [*pasta* resource](https://robinsonlabuzh.github.io/pasta) 
[@Emons2025-pasta], which more comprehensively covers point pattern- and 
lattice-based uni- and multi-variate methods for img- and seq-based ST data.
Notably, *pasta* also includes Python examples and more statistical detail,
e.g., formal definitions and equations for different models and statistics.

- [Voyager](https://pachterlab.github.io/voyager/index.html) offers detailed 
documentation for lattice-based analysis of different data types [@Moses2023-Voyager].

- Point pattern-based content is largely based on the 
[Spatial Point Patterns: Methodology and Applications with R](https://book.spatstat.org/) 
textbook [@Baddeley2015-book-point-patterns].

- @zormpasMappingTranscriptomeRealizing2023 review how methods from 
geographical sciences can be leveraged for the analysis of ST data 
and where potential challenges lie.

- @bianEstimatingTumourImmune2026 discuss spatial statistics methodology 
around a specific biological use case: estimating tumor infiltration.

:::

### References {.unnumbered}

{{< include _stop.qmd >}}
