# Trajectory inference {#sec-ind-trajectory-inference}

{{< include _start.qmd >}}

## Preamble

### Introduction

Trajectory inference (TI) aims to reconstruct dynamic biological processes, such 
as cellular differentiation or disease progression, from snapshot measurements 
of single cells or spatial locations. In single-cell RNA-sequencing (scRNA-seq), 
TI typically relies on the assumption that a population of cells contains 
individuals at different stages of a continuous process. By ordering these cells 
along a path of minimal transcriptional change, we can infer a continuous 
variable known as **pseudotime**, which represents the relative progression of 
each cell along the identified process.

Depending on the underlying biological system, trajectories can exhibit various 
topologies, ranging from simple linear paths to complex branching structures 
(lineages) or even cycles (e.g., the cell cycle). Numerous methods have been 
developed for scRNA-seq data, including `r BiocStyle::Biocpkg("slingshot")` 
[@Street2018-slingshot], `r BiocStyle::Githubpkg("cole-trapnell-lab/monocle3")` 
[@Trapnell2014-monocle; @Cao2019-monocle3], and `PAGA` [@Wolf2019-PAGA].

In the context of spatial transcriptomics (ST), TI provides an opportunity to 
reconcile transcriptional dynamics with physical tissue architecture. Spatially 
aware TI methods leverage the fact that biological transitions often occur across 
contiguous tissue domains, allowing for the reconstruction of more biologically 
grounded developmental paths in what is sometimes called **pseudo-space-time**.

### Dependencies

```{r load-deps, message=FALSE, warning=FALSE}
library(ggplot2)
library(OSTA.data)
library(patchwork)
library(RANN)
library(scrapper)
library(slingshot)
library(SpatialExperiment)
library(SpatialExperimentIO)
```

```{r load-data}
# retrieve dataset from OSF repository
id <- "Xenium_HumanColon_Oliveira"
pa <- OSTA.data_load(id, mol=FALSE)
dir.create(td <- tempfile())
unzip(pa, exdir=td)
(spe <- readXeniumSXE(td, addTx=FALSE))
```

```{r utils}
#| code-fold: true
.plt_xy <- \(spe, col) {
    df <- data.frame(colData(spe), spatialCoords(spe))
    if (col %in% rownames(spe)) df[[col]] <- logcounts(spe)[col, ]
    aes <- if (is.numeric(df[[col]])) {
        theme(
            legend.key.height=unit(1, "lines"), 
            legend.key.width=unit(0.5, "lines"))
    } else {
        list(
            theme(legend.key.size=unit(0, "lines")),
            guides(col=guide_legend(override.aes=list(alpha=1, size=2))))
    }
    ggplot(df, aes(x_centroid, y_centroid, col=.data[[col]])) +
        coord_equal() + theme_void() + aes +
        geom_point(stroke=0, size=1/3)
} 
```

# Trajectory inference

## Non-spatial

```{r area-norm}
# cell area-based normalization
sfs <- (. <- spe$cell_area) / median(.)
spe <- normalizeRnaCounts.se(spe, size.factors=sfs)
```

For this demo, we will *very roughly* subset epithelia (`epi`) by filtering 
for cells with at least 5 counts of PIGR or CEACAM6 transcripts. In reality, 
these should be identified through propoer clustering and annotation, 
or single-cell reference-based label transfer (see @ind-clustering).

```{r sub-epi}
ok <- 
    counts(spe)["PIGR", ] < 2 &
    counts(spe)["CEACAM6", ] > 4
ncol(epi <- spe[, ok])
```

```{r plt-epi-xy, fig.width=8, fig.height=4}
#| code-fold: true
.plt_xy(epi, "AREG") + 
.plt_xy(epi, "CEACAM5") + 
    plot_layout(nrow=1) &
    scale_color_gradientn(colors=pals::jet())
```

Using `r BiocStyle::Biocpkg("slingshot")` [@Street2018-slingshot]

```{r}
# perform principal component analysis (PCA);
# here, using all features due to low plexity
epi <- runPca.se(epi, features=rownames(epi))
```

Next, we perform a quick-clustering in order to identify AREG$-$ cells:

```{r}
epi <- clusterGraph.se(epi, method="leiden", resolution=1)
(go <- names(which.min(by(logcounts(epi)["AREG", ], epi$clusters, mean))))
```

Let's verify this by viewing cluster assignment spatially. Indeed, cluster 
`r use` lies at the base of healthy epithelia (cf. @sec-img-workflow-xenium).

```{r plt-epi-xy, fig.width=4, fig.height=4}
#| code-fold: true
.plt_xy(epi, "clusters") + scale_color_manual(values=unname(pals::trubetskoy()))
```

```{r message=FALSE}
# downsample to at most 10,000 cells
# pet cluster (for runtime reasons!)
res <- slingshot(epi, 
    start.clus=go, 
    reducedDim="PCA", 
    approx_points=200,
    clusterLabels=epi$clusters)
```

```{r}
pt <- "slingPseudotime"
pt <- grep(pt, names(colData(res)))
pt <- as.matrix(colData(res)[pt])
epj$pt <- rowMeans(pt, na.rm=TRUE)
.plt_xy(epj, "pt") + scale_color_gradientn(colors=pals::jet())
```

## Spatially aware {#sec-ind-trajectory-inference-spatial}

While non-spatial methods like `slingshot` can be applied to ST data by using 
spatially-informed embeddings (e.g., from `Banksy` or `SpatialPCA`), several 
specialized frameworks have been developed to explicitly model trajectories 
using both transcriptional similarity and physical proximity:

- **Spatially-aware dimension reduction** methods like `SpatialPCA` 
[@Lu2023-SpatialPCA] and `SpiceMix` [@Chidester2023-SpiceMix] extract 
latent representations that preserve spatial autocorrelation. In principal,
standard TI algorithms can then be applied to these embeddings to produce 
trajectories that are spatially smoothed.

- **Graph-based integration** methods construct cell-cell graphs where edges 
represent both transcriptional similarity and spatial proximity. For example, 
`r BiocStyle::Githubpkg("biomedical-hub/stLearn")` [@Pham2023-stLearn] uses a 
"pseudo-space-time" distance that penalizes transitions between distant points.

- **Optimal transport-based methods** such as `spaTrack` [@Gao2024-spaTrack]
calculate transition costs between cellular states by reconciling expression 
differences with physical distance. This assures that inferred trajectories 
follow physically plausible paths through the tissue.

- **Deep learning** tools such as `SpaceFlow` [@Ren2022-SpaceFlow] use 
spatially regularized graph networks to learn low-dimensional embeddings 
specifically designed for trajectory reconstruction from ST data.

Per usual, the choice of method often depends on the biological question (e.g., 
tracing differentiation across tissue layers vs. identifying migration paths),
as well as spatial resolution and molecular coverage of the platform.

## Appendix

::: {.callout-tip icon="false" title="Further reading"}

- @Cannoodt2016-review-ti review TI in its early days, yet nicely
summarize key aspects of TI modeling, evaluation, and application. 

- @Saelens2019-benchmark-ti provide the most comprehensive benchmark of 
single-cell TI methods to date, including a wide variety of synthetic 
and real datasets, trajectory topologies, and both R and Python tools.

:::

### References {.unnumbered}

{{< include _stop.qmd >}}
