# Differential colocalization {#sec-mult-diff-colocalization}

{{< include _start.qmd >}}

## Preamble

### Introduction

In this section, we will explore tools to assess differential colocalization between conditions with the two packages `r BiocStyle::Biocpkg("spicyR")` [@Canete2022-spicyR] and `r BiocStyle::Biocpkg("spatialFDA")` [@Emons2026-spatialFDA]. Both packages are based the comparison of known spatial statistics functions implemented in the package `r BiocStyle::CRANpkg("spatstat")`.


### Dependencies

```{r, setup, warning=FALSE, message=FALSE}
library(osfr)
library(dplyr)
library(tidyr)
library(refund)
library(spicyR)
library(stringr)
library(ggplot2)
library(patchwork)
library(spatialFDA)
library(SpatialExperiment)
```


### Load data

For this example, we use a dataset on type I diabetes (T1D) [@Damond2019-IMC-diabetes]. This dataset contains three conditions (non-diabetic, onset, and long-duration diabetes), several patients per condition, and several fields-of-view per patient. This hierarchical setup requires the modeler to account for correlation between fields-of-view, for example if they stem from the same patient. Such a setup can be modeled with mixed models (see below).

```{r, load-data-hide, warning=FALSE, eval=FALSE, echo=FALSE}
# load data from full dataset (not run due to computational limitations)
spe <- .loadExample(full = TRUE)
# subset data to smaller example
spe <- spe[, spe$patient_id %in% c(6089, 6180, 6126, 6134, 6228, 6414)]
# remove assays (for smaller object size)
assays(spe) <- list()
```

An assay-free version of the dataset is available through our OSF repository
(used here to reduce runtime as the following analyses do not require assay data):

```{r, load-data, message=FALSE}
# load data from OSF repository (for smaller download)
osf_repo <- osf_retrieve_node("https://osf.io/5n4q3/")
osf_files <- osf_ls_files(osf_repo, 
    path="zzz", n_max=Inf,
    pattern="Damond_noAssays.rds")
dir.create(td <- tempfile())
foo <- osf_download(osf_files, path=td)

# read into R & subset patient IDs (to reduce runtime)
spe <- readRDS(file.path(td, list.files(td, ".rds")))
ids <- c(6089, 6180, 6126, 6134, 6228, 6414)
(spe <- spe[, spe$patient_id %in% ids])
```


## spicyR

First, we show the functionality of `r BiocStyle::Biocpkg("spicyR")`. 
This package condenses the information of a spatial statistics curve 
to a scalar value and models this continuous value as the response 
in a linear (mixed) effects model [@Canete2022-spicyR]. 

By default, `spicyR` chooses the first level of the `condition` 
factor as the base, and the second as the comparison condition;
we first refactor patient stages accordingly:

```{r prep-data}
stages <- c("Non-diabetic", "Onset", "Long-duration")
spe$patient_stage <- factor(spe$patient_stage, stages)
spe$cell_type <- factor(spe$cell_type)
```

To save on runtime, we here skip over some larger `cell_type`s.
By default, `from/to=NULL` will test all pairwise combinations.

```{r, spicyR, message=FALSE, warning=FALSE}
# filter out some larger 'cell_type's
# (for runtime reasons in this demo)
ids <- setdiff(
    levels(spe$cell_type), c("endothelial",
    "alpha", "acinar", "ductal", "unknown"))
.spe <- spe[, spe$cell_type %in% ids]

# perform spatial tests
spicyTestPair <- spicy(
    cells=.spe, 
    condition="patient_stage", 
    imageID="image_name", 
    cellType="cell_type",
    window="square", 
    cores=4)

# extract most significant pairs
head(topPairs(spicyTestPair))
```

The results can be represented as a bubble plot using the `signifPlot` function.

```{r, spicyR-heatmap, fig.width=8, fig.height=6}
signifPlot(spicyTestPair)
```

Here, we can observe that the most significant relationship occurs between 
cytotoxic T cells (Tc) and the beta cells of the pancreas. Tc and beta cells 
appear to be significantly more co-localized in patients with onset diabetes 
compared to non-diabetic patients, indicated by the positive coefficient.

We can use `spicyBoxPlot` to examine a cell type-cell type relationship closely.

```{r, spicyR-boxplot, warning=FALSE, fig.width=5, fig.asp=1}
spicyBoxPlot(spicyTestPair, from="Tc", to="beta") + ylim(c(-50, 50))
```

We can investigate individual images to ensure our results reflect biologically 
meaningful relationships. We can extract the individual co-localization metrics 
from the `spicyTest` object using the `bind` function.

```{r spicyR-metadata, eval=FALSE}
# extract co-localization metrics
cd <- colData(.spe) |>
    as.data.frame() |> 
    distinct(patient_id, patient_stage, image_name) |>
    dplyr::rename(imageID=image_name)
bind(spicyTest) |> 
    # merge with metadata
    merge(cd, by="imageID", all=FALSE) |>
    # select columns of interest
    select(c(imageID, Tc__beta, patient_id, patient_stage)) |>
    filter(imageID %in% c("L29", "O02", "E17", "G17"))
```

We can then visualize the spatial relationship between a pair of 
cell types in the original image using the `plotImage` function.

- The left image (L29) comes from a non-diabetic patient. 
We can see that beta and Tc cells appear to be dispersed, 
which is reflected in the co-localization metric.

- The right image (O02) comes from a patient with early onset diabetes. 
We can see that beta and Tc cells are far more co-localized, 
with Tc cells even infiltrating the pancreatic islets in some cases.

```{r spicyR-plotImage, fig.width=10, fig.height=4.5}
# this is needed due to a bug in 'spicyR' 
# (setting 'cellType' doesn't work)
.spe$cellType <- .spe$cell_type
lapply(c("L29", "O02"), \(i) {
    plotImage(.spe, 
        imageToPlot=i, 
        #cellType="cell_type",
        imageID="image_name",
        from="Tc", to="beta") 
}) |>
    wrap_plots(nrow=1) &
    coord_equal()
```

`spicyR` can also be used with custom distance metrics and to perform 
survival analysis, which are outlined in the [spicyR vignette](
https://www.bioconductor.org/packages/devel/bioc/vignettes/spicyR/inst/doc/spicyR.html).


## spatialFDA

The `r BiocStyle::Biocpkg("spatialFDA")` package computes spatial summaries 
as functions (e.g., Ripley's $K$ function) and compares them using tools from 
functional data analysis (FDA). We will consider the colocalization (spacing) 
of alpha cells in the islet and cytoxic T-cells, which are important in the 
destruction of islet cells in T1D [@Damond2019-IMC-diabetes].

::: {.aside}
This subchapter is adapted from the [spatialFDA vignette](
https://bioconductor.org/packages/devel/bioc/vignettes/
spatialFDA/inst/doc/DiabetesIsletExampleBrief.html).
:::

The first important question is in what range we want to consider interactions. 
We provide a generalisation of a guideline for the maximum radius provided by
@Baddeley2015-book-point-patterns, which suggests to *not* consider any radii 
beyond $1/2$ of the bounding-circle radius of an image, as else effects at the
border of a FOV become too extreme. 

The histogram of all bounding-circle radii across all images and $1/2$ 
of the smallest bin of this histogram as a heuristic are shown below. 

```{r, rMaxHeuristic, results="hide", message = FALSE, warning = FALSE, fig.width=8, fig.height=4}
rMaxHeuristic(spe, subsetby="image_number", marks="cell_type")
```

We will consider a radius of $50 \mu m$ to 
a) be below the heurstic of $91 \mu m$ defined above, and 
b) consider shorter range interactions in our data. 
With a cell diameter of $\sim 10 \mu m$, the `rMax` of 
$50 \mu m$ results in interactions across 5 cell pairs. 

Apart from the radius, we have to define some other parameters as well. 
Important are: 

- `selection` defines which cell types to consider (`NULL` to compare all)
- `fun` specifies the spatial function 
- `marks` specifies cell type annotation in the `colData`
- `correction` specifies which border correction to compare in the functional model
- `transformation` specifies whether to apply a variance-stabilising transformation

The parameters `sample_id, image_id, condition` are important for the statistical inference. 
All other parameters can be specified to achieve better results but there are defaults specified as well. 

```{r, spatialFDA-calc, warning=FALSE}
# relevel to have non-diabetic as the reference category
spe$patient_stage <- relevel(factor(spe$patient_stage), "Non-diabetic")

# run the spatial statistics inference
resLs <- crossSpatialInference(
    spe, 
    selection=c("alpha", "Tc", "Th"),
    fun="Gcross", 
    marks="cell_type",
    rSeq=seq(0, 50, l=50), 
    correction="rs",
    transformation="Fisher",
    eps=1e-3,
    delta="minNnDist",
    family=mgcv::scat(link="log"),
    sample_id="patient_id",
    image_id="image_name", 
    condition="patient_stage",
    algorithm="bam",
    verbose=FALSE)
```

The convenience function `crossSpatialInference` first calculates the spatial 
statistics functions, and then compares these with a functional additive mixed 
model across all cell types in the `selection` 
[@Baddeley2015-book-point-patterns; @Baddeley2005-spatstat; 
@Scheipl2015-additive; @Scheipl2016-generalized; @Goldsmith2024-refund-package]. 

The result is a list of lists containing the results across all cell types:

```{r spatialFDA-resLs1}
# list with one element per cell type
names(resLs)
```

Results for each cell type pair include the transformed functions (`metricRes`), 
the untransformed functions (`metricResRaw`), the design matrix used to fit the 
model (`designmat`), the actual functional model (`mdl`) and some QC metrics of 
the statistical inference (`curveFittingQC`):

```{r spatialFDA-resLs2}
# list of results for first cell type
names(resLs[[1]])
```

We can plot the overall picture of the interactions by performing an $F$-test 
on all functional models and plotting both the $p$-value of the tests as well 
as the mean coefficient over the domain $r$.

```{r, spatialFDA-bubbleplot, results="hide", fig.width=8, fig.height=4}
coefs <- c(
    "Onset"="conditionOnset(x)", 
    "Long-Duration"="conditionLong_duration(x)")
plotCrossHeatmap(resLs, 
    coefficientsToPlot=coefs, QCThreshold=0, QCMetric="medianMinIntensity") + 
    guides(shape="none") + facet_wrap(~factor(coefficient, coefs, names(coefs)))
```

This plot highlights an increased co-localisation of T-cell types with 
pancreatic $\alpha$ cells in Onset diabetes but not in Long-duration diabetes. 
This is an interesting overall finding and we will focus on the cell type pair 
$\alpha$-cytotoxic T cells in more detail below.

First, we will plot the functional box plot of the 
transformed spatial statistics functions [@Sun2011-boxplots].
Here, the black line indicates the median curve, the purple region 
is the 50% central region, and the red lines indicate outliers.

```{r, spatialFDA-plot, warning=FALSE, fig.width=4, fig.height=4}
#| layout-ncol: 3
# extract the metric dataframe
res <- resLs$alpha_Tc
metricRes <- res$metricRes

# make unique identifiers
metricRes$ID <- with(metricRes, paste(sep="x", 
    patient_stage, patient_id, image_name))

# functional boxplot of spatial statistics curves
collector <- plotFbPlot(metricRes, "r", "rs", "patient_stage")
```

From these plots, we note that the functions from 'Onset' 
patients seem to qualitatively have a higher median curve 
than those from 'Non-diabetic' or 'Long-duration' patients.

Another exploratory technique is functional PCA (FPCA). 
FPCA is similar to standard PCA in that it decomposes 
the functional data into the main modes of variation. 
Here, we decompose the functional data into eigenfunctions 
and can plot the first two principal components as a biplot 
[@Goldsmith2024-refund-package].

```{r, spatialFDA-fPCA, fig.width=8, fig.height=3}
# prepare 'data.frame' from 'calcMetricRes' 
# to be in the correct format for FDA
dat <- prepData(metricRes, "r", "rs")

# create meta info of the IDs
splitData <- dat$ID |> 
    str_replace("-","_") |> 
    str_split_fixed("x", 3) |> 
    data.frame(stringsAsFactors=TRUE) |> 
    setNames(c("condition", "patient_id", "imageId")) |> 
    mutate(condition=relevel(condition,"Non_diabetic"))

dat <- cbind(dat, splitData) # join with results
dat <- drop_na(dat)          # drop rows with NA

# calculate FPCA & visualize biplot
pca <- functionalPCA(dat=dat, r=unique(metricRes$r), pve=0.995)
plotFpca(dat=dat, res=pca, colourby="condition") + labs(color="condition")
```

From the FPCA, we note that the variability in curves 
from 'Onset' patients is larger than the other two groups.

Next, we will show how FDA can be used to make inferences from functional spatial 
summaries. We use a generalized additive mixed model with a functional response 
[@Scheipl2015-additive; @Scheipl2016-generalized; @Goldsmith2024-refund-package]. 
In such models, some parameters are functional and some are constants (see below).

```{r, spatialFDA-fGAM, results="hide", fig.width=6, fig.height=5}
summary(mdl <- res$mdl)
mm <- res$designmat
lapply(colnames(mm), \(x) {
    i <- mdl$coefficients[[1]] 
    plotMdl(mdl, predictor=x, shift=i)
}) |> wrap_plots(nrow=3, axes="collect")
```

With the functional GAM, we see not only the effect (e.g., increased 
colocalization of $\alpha$ and cytotoxic T cells in 'Onset' patients compared
to 'Non-diabetic' patients) but also the strength of the effect at each scale.

The Q-Q plot indicates a reasonable model fit, although not perfect.

```{r, spatialFDA-qq, fig.width=5}
res <- mdl$residuals
qqnorm(res, pch=16)
qqline(res, col="red")
```

In summary, we have seen that there is increased co-localisation of T-cell types
to the pancreatic islets early in the disease and that this recruitment is absent
later in the disease. This increased co-localisation in onset diabetes is happening
at shorter scales around $\alpha$ cells indicating that effects close to the 
pancreatic islets are more affected than those further away.

## Appendix

::: {.callout-tip icon="false" title="Further reading"}

- [mxFDA](https://github.com/julia-wrobel/mxfda/) [@Wrobel2024-mxfda]
is an R toolkit for functional data analysis (FDA) of replicated single-cell
spatial data, and associates spatial relationships to patient outcomes.

- [CRAWDAD](https://github.com/jefworks-lab/crawdad)[@Peixoto2025-CRAWDAD]
is an R package for multi-scale characterization of spatial relationships
between cell subpopulations, including comparison across multiple samples.

- `r BiocStyle::Biocpkg("smoppix")`[@Hawinkel2026-smoppix] 
is an R/Bioconductor package that provides univariate and bivariate tests 
for spatial patterns (e.g., gradients and vicinity to cell centroid/edge),
including support for nested design and automatic parameter calibration.

:::

### References {.unnumbered}

{{< include _stop.qmd >}}
