# Structure-based analysis {#sec-mult-structure-based}

{{< include _start.qmd >}}

## Preamble

### Introduction

Spatial omics data allow us to quantify various features that are related to tissue architecture. This allows us to perform comparisons of anatomical structure-derived features. We will refer to this type of analysis as **structure-based analysis**.

We will use the `r BiocStyle::Biocpkg("sosta")` package [@Gunz2025-sosta] to reconstruct and quantify pancreatic islets from human donors with different stages of type 1 diabetes (T1D) and healthy controls [@Damond2019-IMC-diabetes]. We will then compare some of the features between conditions.

This chapter is closely related to the spatial statistics and neighborhood concepts introduced in @sec-ind-spatial-statistics and @sec-img-neighborhood-analysis, but it summarizes those spatial patterns as structure-level geometric features before comparing groups of samples.

### Dependencies

```{r load-deps, message=FALSE, warning=FALSE}
library(sf)
library(osfr)
library(lme4)
library(dplyr)
library(tidyr)
library(sosta)
library(ggplot2)
library(lmerTest)
library(ggfortify)
library(ExperimentHub)
library(SpatialExperiment)
# set seed for random number generation
# in order to make results reproducible
set.seed(8048)
```


### Load data

<!--
We will load the data directly from `ExperimentHub` and convert it into a `SpatialExperiment` object. To reduce runtime, we will subset the data to 15 images per patient.

```{r load-data-structs, message=FALSE, eval=FALSE, echo=FALSE}
# load data from ExperimentHub and convert to SpatialExperiment
# (not run due to computational limitations)

eh <- ExperimentHub()
oid <- names(eh[eh$title == "Damond_2019_Pancreas - sce - v1 - full"])

sce <- eh[[oid]]

spe <- toSpatialExperiment(
  sce, 
  sample_id = "image_name", 
  spatialCoordsNames = c("cell_x", "cell_y"))

# remove assays (for smaller object size)
assays(spe) <- list()
```
-->

An assay-free version of the dataset is available through our OSF repository
(used here to reduce runtime as the following analyses do not require assay data):

```{r load-data}
# load data from OSF repository (for smaller download)
osf_repo <- osf_retrieve_node("https://osf.io/5n4q3/")
osf_files <- osf_ls_files(osf_repo, 
    path="zzz", n_max=Inf,
    pattern="Damond_noAssays.rds")
dir.create(td <- tempfile())
foo <- osf_download(osf_files, path=td)
spe <- readRDS(file.path(td, list.files(td, ".rds")))
table(spe$patient_id, spe$patient_stage)
```


## Visualize data

First, we plot a few selected fields of view (FOVs).

```{r plot-xy1, fig.width=7, fig.height=4.5}
# subset to 4 randomly selected FOVs
fov <- sample(unique(spe$image_name), 4)
sub <- spe[, spe$image_name %in% fov]
df <- data.frame(colData(sub), spatialCoords(sub))
# visualize annotations
ggplot(df, aes(cell_x, cell_y, color=cell_category)) + 
    geom_point(size=0.4) + facet_wrap(~image_name, ncol=2) + 
    scale_color_manual(values=unname(pals::okabe(n=5))) +
    guides(col=guide_legend(override.aes=list(size=2))) +
    coord_equal() + theme_classic() 
```


## Islet reconstruction

Next, we will reconstruct or segment the individual islets using a density-based approach implemented in `r BiocStyle::Biocpkg("sosta")`.

```{r est-params}
# use 'nImages' randomly samples 
# images for parameter estimation
est <- estimateReconstructionParametersSPE(
    spe, 
    marks="cell_category", 
    imageCol="image_name", 
    markSelect="islet", 
    nImages=10, 
    nCores=4, 
    plotHist=FALSE)
# get parameter estimates for ...
th <- mean(est$thres) # threshold
bw <- mean(est$bndw)  # bandwidth
```

::: {.callout-note collapse="true" title="Density-based reconstruction"}

The `r BiocStyle::Biocpkg("sosta")` package uses a reconstruction method based on the point pattern *density* of the islet cells. This method needs two parameters: 

- a bandwidth that is used for intensity profile estimation, and 
- a threshold based on which the islets are reconstructed.

```{r plot-img, fig.width=8, fig.height=3}
shapeIntensityImage(
    spe, 
    marks="cell_category", 
    imageCol="image_name", 
    imageId=fov[4], 
    markSelect="islet")
```

Shown on the left is a density (pixel-level) image, and on the right a histogram of the intensity values. The method selects every pixel above a certain threshold for reconstruction. The smoothing bandwidth that generates the image is estimated using cross-validation (see function `bw.diggle()` of the `r BiocStyle::CRANpkg("spatstat.explore")` package from CRAN).

The threshold is estimated by taking the mean between the two modes of the (truncated) pixel intensity distribution. The function `estimateReconstructionParametersSPE()` repeats the estimation of these parameters for a selection of images in the dataset.

To make the reconstruction comparable between images we then use a fixed set of parameters for all images, in our case the mean of the individual parameters for a subset of the dataset.

:::

```{r get-shapes}
allIslets <- reconstructShapeDensitySPE(
    spe, 
    marks="cell_category", 
    imageCol="image_name", 
    markSelect="islet", 
    bndw=bw, 
    thres=th, 
    nCores=4)
```

Now we can inspect the reconstruction in the sample images.

```{r plot-xy2, fig.width=7, fig.height=4.5, message=FALSE}
# subset to selected FOVs
subIslets <- allIslets[allIslets$image_name %in% fov, ]
# visualize annotations 
ggplot(df, aes(cell_x, cell_y, color=cell_category)) + 
    geom_point(size=0.4) + facet_wrap(~image_name, ncol=2) + 
    scale_color_manual(values=unname(pals::okabe(n=5))) +
    guides(col=guide_legend(override.aes=list(size=2))) +
    coord_equal() + theme_classic() +
    geom_sf( # geom for structure outlines
        data=subIslets, inherit.aes=FALSE, 
        color="red", fill=NA, linewidth=1)
```

The `allIslets` object is a [simple feature collection](https://r-spatial.github.io/sf/articles/sf1.html) which contains polygons (`<GEOMETRY>` column), a structure identifier (`structID`), and the image identifier (`image_name`). We will add some patient metadata to the object.

```{r join-meta}
# factor levels for 'patient_stage'
lv <- c("Non-diabetic", "Onset", "Long-duration")
# cell metadata columns to keep
colsKeep <- c(
    "sample_id", "image_name",
    "patient_disease_duration", 
    "patient_id", "patient_stage", 
    "patient_age", "patient_gender",
    "tissue_slide", "tissue_region") 
patientData <- colData(spe) |> 
    as_tibble() |> 
    # keep selected columns
    select(all_of(colsKeep)) |> 
    # refactor patient IDs & stages
    group_by(image_name) |> 
    mutate_at("patient_id", factor) |>
    mutate_at("patient_stage", factor, lv) |>
    # keep only unique combinations
    unique() 
# join with results from reconstruction
allIslets <- left_join(allIslets, patientData, by="image_name")
```


## Quantification of geometric features

Now we can proceed with quantification of geometric aspects 
of the islets, and combine them with patient information.

```{r quant}
isletMetrics <- totalShapeMetrics(allIslets)
allIslets <- cbind(allIslets, t(isletMetrics))
```

PCA can give us an overview of the different features; here, each dot represents one structure.

```{r plot-pca, warning=FALSE, fig.width=7, fig.height=5}
#| code-fold: true
pca <- prcomp(t(isletMetrics), scale.=TRUE)
autoplot(pca, 
    x=1, y=2, 
    data=allIslets,
    color="patient_stage", 
    size=2, 
    loadings=TRUE, 
    loadings.colour="steelblue3", 
    loadings.label=TRUE, 
    loadings.label.size=3, 
    loadings.label.repel=TRUE, 
    loadings.label.colour="black") +
    theme_bw() + coord_fixed() + 
    scale_color_manual(values=unname(pals::tol(n=3)))
```

```{r plot-mets1, fig.width=6, fig.height=4}
# wrangling
df <- allIslets |> 
    st_drop_geometry() |> 
    select(patient_stage, rownames(isletMetrics)) |> 
    pivot_longer(-patient_stage) |> 
    filter(name %in% c("Area", "Compactness", "Curl"))
# visualization
ggplot(df, aes(patient_stage, value, fill=patient_stage)) + 
    geom_violin() + geom_boxplot(aes(fill=NULL), width=0.3) + 
    scale_fill_manual(values=unname(pals::tol(n=3))) +
    scale_x_discrete(guide=guide_axis(n.dodge=2)) + 
    facet_wrap(~name, scales="free") + 
    guides(fill="none") + theme_bw()
```

Let's focus on the area of the islets and facet by stage to look at patient variability. As the distribution is very skewed, we will use a 1/4-power transformation on the area of the islets. The transformation was chosen after inspection of the model diagnostics (see below).

```{r plot-mets2, fig.width=6, fig.height=4}
# wrangling
df <- allIslets |> 
    st_drop_geometry() |> 
    select(patient_stage, patient_id, rownames(isletMetrics)) |> 
    pivot_longer(-c(patient_stage, patient_id)) |> 
    filter(name %in% c("Area"))
# visualization
ggplot(df, aes(patient_id, value^(1/4), fill=patient_stage)) + 
    geom_violin() + geom_boxplot(fill=NA, width=0.3) + 
    scale_fill_manual(values=unname(pals::tol(n=3))) +
    facet_wrap(~patient_stage, scales="free_x") + 
    geom_jitter(size=0.5) + ylab("Area") +
    guides(fill="none") + theme_bw()
```


## Between-sample comparison

An important aspect to note is that the individual structure metrics are not independent measurements, since there are generally repeated measurements (i.e., multiple islets) per slide and per patient. Therefore, we need to account for this correlation between measurements.

To account for this, we will use mixed linear models with random effects for the patient and the individual slides (image name). The `r BiocStyle::CRANpkg("lme4")` package will be used for fitting linear mixed effects models [@Bates2015-lme4] and `r BiocStyle::CRANpkg("lmerTest")` for p-value calculations [@Kuznetsova2017-lmerTest].

::: {style="max-height:500px; overflow-y: auto;"}

```{r mod}
f <- Area^(1/4) ~ patient_stage + (1|patient_id) + (1|image_name)
summary(mod <- lmer(formula=f, data=allIslets))
```

:::

From the fixed effects section in the model `summary` above, we note that there is a statistically significant difference in the transformed islet area of long-duration patients compared to non-diabetic patients, while the effect for onset patients is not significant at the 5\% level. This result accounts for correlation at the patient and image level as modeled by random intercepts. The transformation (1/4-power) was chosen after inspection of the residual behavior in the model diagnostics for this dataset and feature of interest specifically.

::: {.callout-note collapse="true" title="Model diagnostics"}

:::: {.panel-tabset}

### Residuals vs. fitted

```{r plot-resfit1, fig.width=6}
res <- resid(., scaled=TRUE) ~ fitted(.)
plot(mod, res, 
    pch=12, abline=0, 
    xlab="Fitted values", 
    ylab="Standardised residuals",
    col=allIslets$patient_id) 
```

### Residuals vs. fitted per sample

```{r plot-resfit2, fig.width=7}
res <- resid(., scaled=TRUE) ~ fitted(.) | patient_id
plot(mod, res, 
    pch=12, abline=0, 
    xlab="Fitted values", 
    ylab="Standardised residuals")
```

### QQ plot

```{r plot-qq, fig.width=5}
qqnorm(resid(mod), pch=16)
qqline(resid(mod))
```

::::

:::

## Appendix

::: {.callout-tip icon="false" title="Further reading"}

- This chapter is based on the `r BiocStyle::Biocpkg("sosta")` vignette "2. Reconstruction and analysis of pancreatic islets from IMC data"; more detail is provided in @Gunz2025-sosta.

- The Bioconductor packages `r BiocStyle::Biocpkg("SPIAT")` and `r BiocStyle::Biocpkg("imcRtools")` offer other functions to quantify features on the anatomical structure level. These could also be input for mixed effect modeling as illustrated above.

- The packages `r BiocStyle::CRANpkg("vegan")` and `r BiocStyle::Biocpkg("poem")` are useful when calculating diversity metrics within structures and/or border regions. To identify border regions and how to overlap anatomical regions with sample coordinates, please have a look at the `r BiocStyle::Biocpkg("sosta")` vignette.

- The Python packages [GRIDGENE](https://demirandalab-gridgen.readthedocs.io) [@Sequeira2025-GRIDGENE] and [CellCharter](https://cellcharter.readthedocs.io) [@Varrone2024-CellCharter] provide complementary approaches for structure-based spatial omics analysis. `GRIDGENE` focuses on analyses based on transcript density, whereas `CellCharter` offers methods for spatial clustering, reconstruction of spatial structures, and quantification of their geometric properties.

:::

### References {.unnumbered}

{{< include _stop.qmd >}}
