# Data import and structure {#sec-seq-data-import-structure}


## Introduction

The following chapters provide examples demonstrating methods for analysis steps for sequencing-based spatial transcriptomics data. Most of these examples rely on datasets formatted as `SpatialExperiment` objects (see @sec-bkg-data-infrastructure).


## Dataset

Here, we import a 10x Genomics Visium dataset that will be used in several of the following chapters. This dataset has previously been preprocessed using data preprocessing procedures with tools outside R and saved in `SpatialExperiment` format, and is available for download from the [STexampleData](https://bioconductor.org/packages/STexampleData) package.

This dataset consists of one sample (Visium capture area) from one donor, consisting of postmortem human brain tissue from the dorsolateral prefrontal cortex (DLPFC) brain region, measured with the 10x Genomics Visium platform. The dataset is described in @Maynard2021.


## Import data from STexampleData package

Download and import the dataset in `SpatialExperiment` format from the [STexampleData](https://bioconductor.org/packages/STexampleData) package.

```{r, message=FALSE}
library(SpatialExperiment)
library(STexampleData)
```

```{r}
# import data
spe <- Visium_humanDLPFC()
# temporary fix
imgData(spe) <- NULL
```

We also save the data object in `.rds` format for re-use within later chapters, to speed up the build time of the book.

```{r}
# save data
saveRDS(spe, file = "seq-spe_load.rds")
```


## Import data from raw files

Alternatively, reader functions from several packages can be used to import data from raw files and create a `SpatialExperiment` object, including:

- The `read10xVisium()` function from the `SpatialExperiment` package can be used to import Visium data from raw data files from the 10x Genomics Space Ranger pipeline
- The [VisiumIO](http://bioconductor.org/packages/VisiumIO) and [XeniumIO](https://bioconductor.org/packages/XeniumIO) packages provide functions to import Visium and Xenium data, respectively
- The [SpatialExperimentIO](http://bioconductor.org/packages/SpatialExperimentIO) package provides functions to import imaging-based spatial transcriptomics data (Xenium, CosMx, MERSCOPE, and STARmapPLUS platforms) from raw data files (see also @sec-img-introduction)


## SpatialExperiment data structure

This section illustrates the structure of the `SpatialExperiment` object loaded above. For more details on the `SpatialExperiment` data structure, see @sec-bkg-data-infrastructure.

```{r}
# show object
spe

# number of genes (rows) and spots (columns)
dim(spe)

# names of 'assays'
assayNames(spe)

# row (gene) data
head(rowData(spe))

# column (spot) data
head(colData(spe))

# spatial coordinates
head(spatialCoords(spe))

# image data (optional)
imgData(spe)
```


## Appendix

### References {.unnumbered}
